---
title: "Chemical co-occurrence networks -- MAINLINE"
subtitle: "Everything that produces Tables 1-4 and Figures 1-4. Method justification, alternative community detection, cross-basin work, the null hierarchy and the SGH test live in the S1-S5 companions."
author: "Robert Neugebauer (vRN)"
date: "14.09.2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 11, fig.height = 5)
```

<!-- Change log and rationale carried over verbatim from the monolithic script. -->
<!-- =====================================================================
     vPI2 — WHAT CHANGED RELATIVE TO THE ORIGINAL SCRIPT
     (the original Clean_Workflow_Updated_No_CLR_vPI.Rmd is left untouched)

     1. WGCNA soft-threshold floor now scales with sample size for a SIGNED
        network (min_power_signed) instead of a fixed floor of 6.
     2. glasso is the PRIMARY weighted network (partial correlations); MB is
        kept only as an edge-set robustness check (ARI / Jaccard). WGCNA is
        kept for MODULES (exposure mixtures), not for edge/density comparison.
     3. Null model is now WEIGHTED (ER structure + resampled observed weights)
        so observed and null modularity are on the same scale.
     4. Prevalence filter is a single parameter (PREV_THRESH) plus a diagnostic
        section to justify its value rather than assuming 10 %.
     5. NEW river-basin-specific networks (all sites of a basin, n >= 20).
     6. NEW absolute-risk-driver overlay after Inostroza et al. (2024):
        RQ (= TU) >= 0.02 at >= 1 site. Structural centrality ("sentinel")
        and risk-driver status are treated as two INDEPENDENT node attributes.
     Every changed / new block below is tagged  ## [vPI2].

     vPI3 — WHAT CHANGED RELATIVE TO vPI2
     7. Non-detect substitution instead of a flat pseudocount: true zeros (MEC = 0, i.e.
        below MDL) are replaced with a compound-specific MDL/2 (raw_data/MDL_CECs_summary.csv,
        same lab/method pooled across projects/years, lowest reported MDL per compound;
        matched by InChIKey, then by name), for both concentrations and the equivalent
        MDL/2-via-EC10 for toxic units. Detected values are left exactly as measured --
        vPI2 added its pseudocount (+1 ng/L, or 0.5*min for TU) to EVERY cell, which
        inflated real detections and, for TU, would have distorted the absolute-risk-driver
        threshold. Compounds without a matched MDL fall back to the vPI2 default constant.
        Every changed block is tagged  ## [vPI3].
     8. Whole Basin (Rhine+Danube) pooled unit removed (was conflating two unrelated
        basins into one artificial group); Rhine and Danube are now only ever run
        independently, either as sections (`groups`) or as whole basins (`basin_units`).
===================================================================== -->

<!-- =====================================================================
     vRN — WHAT CHANGED RELATIVE TO vPI3 (Robert, 04.08.2026)
     Every new/changed block is tagged  ## [vRN].
     ADDITIONS:
     A. Rarefied CROSS-BASIN comparison + module-sharing (the gap the report flags):
        every basin subsampled to a common n, glasso refit R times, topology averaged,
        and a pairwise module-sharing index (do the same compounds co-cluster across basins?).
     B. Latitude/longitude UPPER/MIDDLE/LOWER sections incl. the ELBE
        (Elbe & Rhine split by latitude, Danube by longitude — it flows W->E; upstream
        end is oriented correctly per basin). Run through the same run_unit() pipeline.
     C. Risk drivers: added a  pct_sumTU  column (share of the unit's total toxic
        pressure), matching Pedro's "% of the calculated risk" phrasing, plus an
        MDL/2 vs MDL vs drop-zeros SENSITIVITY check on a representative unit.
===================================================================== -->

# Rationale (read me first)

## Why partial correlations, and why glasso {#rationale-partial}

Two chemicals can seem to "go together" across sites for two very different
reasons, and it matters which one we are seeing.

The simplest measure is a **marginal** correlation: do the two compounds rise
and fall together across sites, yes or no? The problem is that it does not ask
*why*. The strongest reason two chemicals move together in a river is usually
not a real link between them at all — it is that some sites are simply more
polluted than others. A site just below a large city or a wastewater outfall
has high levels of almost everything; a clean upstream site has low levels of
almost everything; and river flow dilutes everything at once. So pick almost any
two compounds and they will appear correlated, just because both are being
carried up and down by this overall pollution-and-dilution level. A marginal
correlation counts that as a connection, even though the two chemicals have
nothing specifically to do with each other.

A **partial** correlation asks the stricter question: once we account for what
all the other measured compounds already tell us, do these two *still* move
together? If their apparent link was only the shared pollution level (which
every other compound also follows), that shared level already explains it,
nothing is left over, and no edge is drawn — which is what we want. But if the
two share something specific that the general pollution level does not explain —
the same individual source, one compound being the breakdown product of the
other, or both simply behaving the same way in water — then they keep moving
together even after everything else is accounted for, and the edge survives.
Those surviving edges are the **direct** associations worth interpreting.

Computing a partial correlation means, in effect, holding all the other
compounds constant for every pair. With many more compounds than sites this
cannot be done the ordinary way — there is not enough data. The graphical lasso
(**glasso**) is the tool that does it stably: it estimates all these direct
links at once and, because the data are limited, stays conservative — forcing
the weak, uncertain links to exactly zero and keeping only the ones strong
enough to trust. That is why the glasso network is sparse and why its edges
carry meaning, rather than being a dense web produced by the shared pollution
gradient.

Because our ecological interpretation rests on these direct links, we estimate the
network from partial correlations. With more compounds than sites (p > n) the
partial-correlation matrix cannot be obtained by ordinary matrix inversion; the
graphical lasso (glasso) solves this by estimating a **regularised (sparse)**
partial-correlation matrix, keeping only edges strong enough to survive
penalisation. This is why glasso, not a plain correlation network, is our primary
tool. Meinshausen–Bühlmann (MB) neighbourhood selection estimates the same graph
by a different route; we keep it not as a second network but as a **robustness
check** — the agreement between glasso and MB edges (ARI / Jaccard) tells us
whether an edge reflects the data or one algorithm's penalty.

**Prevalence filtering and glasso sparsification are orthogonal.** The two steps are
sometimes read as pulling against each other -- one discards compounds, the other discards
links, so why do both? They act on different objects. The prevalence filter operates on the
**node set**: it decides which compounds are admitted as variables at all, on the grounds
that a compound detected at a handful of sites carries too little support for any
correlation involving it to be estimated reliably (Herren & McMahon 2017, ISME J 11:2426,
the origin of the practice; Hernandez et al. 2021, ISME J 15:1722, the SGH application this
pipeline follows). The glasso penalty operates on the **edge set**: given the admitted
variables, it decides which conditional dependencies are distinguishable from zero. Neither
substitutes for the other, and applying both is not double sparsification of the same
quantity. The one real interaction is that the filter sets p, and p relative to n governs
how hard the penalised problem is -- a stability and computation consideration (see the
prevalence diagnostic and A44), not a conceptual conflict.

In this dataset the filter also does a job the microbiome literature never needed it for.
Our zeros are left-censored non-detections replaced by an MDL/2 floor, so two
mostly-undetected compounds can correlate through shared non-detection alone. The
prevalence filter is the control for that artefact, and the effect is measurable: with no
filter, more than a quarter of edges rest on fewer than four joint *detections*; at the
10 % filter, not a single edge does.

## How glasso and WGCNA complement each other {#rationale-wgcna}

glasso and WGCNA answer two complementary questions and are not competing versions
of the same network. glasso is a **conditional-dependence** method: each edge is a
direct, partial-correlation link, best suited to tracing pairwise mechanisms —
which two compounds share a source, a metabolic pathway, or a partitioning
behaviour. WGCNA is a **marginal, module-oriented** method: it does not remove
shared drivers but instead groups compounds into modules of mutually co-varying
substances. Those modules are the natural unit for an **exposure-mixture**
interpretation — chemicals that rise and fall together across sites and therefore
co-expose aquatic organisms. In short, glasso resolves the **wiring** between
individual compounds (sources, direct links); WGCNA delineates the **blocks** of
the network (co-occurring mixtures). Because the two rest on different definitions
of association (conditional vs. soft-thresholded marginal), we do not compare their
edge counts or densities directly — only method-agnostic outcomes: which compounds
cluster together, and which act as hubs.

## Why modularity, what a module means, and what it cannot do {#rationale-modularity}

Each network is partitioned by greedy modularity maximisation (`cluster_fast_greedy`).
Three things need saying about that choice, because the method is more often used than
justified.

**What modularity is.** Newman (2006, PNAS 103:8577) defines Q as the number of edges
falling inside groups minus the number expected in an equivalent randomised network, and
finds the partition by *optimising* it. It is a quality function and a data-analysis
technique, not a hypothesis test: the paper proposes no significance procedure, and in most
applied work none is used. The expected-edge term is k_i*k_j/2m -- a configuration-model
expectation computed from the observed degrees -- so Q already carries a degree-preserving
null inside it. That is why any external ensemble it is compared against must also be
degree-preserving (see the Table 3 caption).

**What a module is taken to mean.** In the microbial-network literature a module is read as
either a group of taxa interacting more with each other than with the rest, or a group
sharing a niche distinct from other groups (Faust & Raes 2012, Nat Rev Microbiol 10:538;
Hernandez et al. 2021; Hou et al. 2024). The hypothesis attached is one of stability:
modular organisation insulates groups from disturbance, so modularity is predicted to
decline along a stress gradient. That prediction is what makes modularity the relevant
statistic for a Stress Gradient Hypothesis test at all.

**What it means here.** Neither reading transfers directly, because compounds do not
interact and occupy no niche. An edge here is co-variation of concentration across sites
after conditioning, so a module is a set of compounds whose concentrations move together
across the river network -- generated by shared emission source, shared transport and
degradation behaviour, or shared application pattern. Which of these dominates is settled
empirically rather than assumed: an edge-level nested model over roughly 18 500 compound
pairs per basin finds **emission source overwhelming (p from 5.0e-04 to 1.7e-60) and mode
of action adding nothing on top of it**. Modules in this dataset are emission-source
compartments. This is also where the SGH transfer needs stating openly: what declines with
stress in Hernandez et al. is niche and interaction compartmentalisation, whereas what this
pipeline measures is source compartmentalisation.

**Known pathologies, and how they are handled.** Modularity maximisation has three that no
choice of random-graph null repairs. Random graphs attain high Q by chance (Guimera et al.
2004, Phys Rev E 70:025101). The objective cannot resolve modules smaller than about
sqrt(2m) edges, so genuinely distinct small groups are merged -- the resolution limit
(Fortunato & Barthelemy 2007, PNAS 104:36). And the Q landscape is degenerate, with
exponentially many near-optimal partitions of very different composition, so a reported
partition is one draw from a plateau rather than a unique answer (Good, de Montjoye &
Clauset 2010, Phys Rev E 81:046106). Three responses are built into this pipeline: the null
test uses degree-preserving ensembles rather than ER; the partition is validated against
independently documented chemistry rather than trusted on its Q alone; and a generative
alternative that does not use modularity at all (SBM / DC-SBM, see #sbm-blocks) is fitted
as a check. In consequence, module-level claims here are made at the level of block
composition -- small modules and the assignment of individual borderline compounds should
be read with the resolution limit and the degeneracy in mind.

## Concentration vs. toxic units: structure vs. risk {#rationale-tu}

The network is built from measured concentrations, because co-occurrence and
shared-source structure are properties of how much of each compound is present and
how it co-varies across sites — questions of emission and transport, not toxicity.
Expressing the same data as toxic units (TU = MEC / EC10) rescales each compound by
its own potency, which is what we want for judging **risk** — but it does not change
the network's wiring: dividing each compound by a constant is, after the
log-transform, only a per-compound shift, and correlations are invariant to such
shifts (empirically r ≈ 0.997 between the two correlation structures). A TU network
is therefore not independent evidence about co-occurrence. Instead, each compound
carries **two independent node attributes** that we are careful not to conflate.
The first is **structural centrality** — how strongly and widely a compound is
connected in the co-occurrence network. A highly central compound is a **sentinel**:
its presence signals the wider mixture it travels with and it may mark a shared
source or transport pathway, but high connectivity says nothing about how toxic or
how concentrated it is. The second is toxicological: whether the compound is an
**absolute risk driver**, defined after Inostroza et al. (2024) as an individual
risk quotient RQ ≥ 0.02 at least at one site — a compound that on its own reaches an
appreciable share of the acceptable mixture risk somewhere in the unit. Connectivity
and risk are **orthogonal**: a compound can be a well-connected sentinel that is
toxicologically unimportant, or a poorly connected, peripheral compound that is
nonetheless a major risk driver at a single hotspot — the latter being exactly the
case a co-occurrence network would overlook if read on its own. We therefore compute
the two attributes separately and overlay the risk-driver status onto the
concentration network, so the analysis asks **where the risk drivers actually sit**
rather than assuming the most connected nodes are the ones that matter for the
organism. (Note: here TU = MEC / EC10_GRO is a *simplified* RQ — single growth
endpoint, no assessment factor — so it approximates rather than exactly reproduces
the paper's RQ_MST; the 0.02 threshold and per-site logic follow the paper.)

## Why a river-basin-specific layer, and how to read it {#rationale-basin}

The country-based groups resolve chemical co-occurrence at a relatively local scale,
where the concentrations separating one site from the next are governed mainly by
proximity to individual inputs; there, an edge can reasonably be read as a shared
point source or a common sub-catchment. In the river-basin-specific networks we
deliberately change scale, pooling every site of a basin regardless of country. At
this scale the dominant variation between sites is no longer which outfall is nearby
but the large-scale gradients of the basin — the upstream-to-downstream progression,
urban versus agricultural land use, flow and dilution, and season. An edge therefore
no longer points to a single shared pipe but to a shared **basin-scale behaviour**:
compounds that travel, persist and accumulate together along the same environmental
and land-use gradients. We keep both scales on purpose. Associations that survive
from the local groups up to the whole basin are robust, specific co-behaviours;
associations that appear only at basin scale are more likely driven by the common
upstream–downstream / dilution gradient. Because glasso conditions each compound pair
on all other compounds, it partially removes that shared gradient and is our
preferred estimator at this scale; the marginal (WGCNA) view will, by contrast, be
more strongly shaped by the basin gradient itself.


# Setup, data and functions

```{r split-source, message=FALSE}
## Everything below assumes the shared setup/data/functions are loaded.
## Repo-relative: knitr sets the working directory to this .Rmd file's folder.
sp <- file.path("..", "R")
source(file.path(sp, "00_setup.R"))
source(file.path(sp, "01_data.R"))
source(file.path(sp, "02_functions.R"))
```


## Writing summary tables etc.

```{r summary-chunk}

run_unit <- function(kind, unit, rep.num = STARS_REPNUM, n_null = 200, scale_tag = "group") {
  kind_label <- if (kind == "conc") "Concentrations (ng/L)" else "Toxic units (TRIDENT EC10)"
  outdir <- file.path(outroot, if (kind == "conc") "concentration" else "toxic_unit", unit$safe)
  dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

  pu <- prep_unit(kind, unit$site_ids)
  if (pu$n_comp < 5 || pu$n_sites < 4) { message("skipping ", unit$label, " (too small)"); return(NULL) }

  gl <- fit_glasso(pu$Xlog, rep.num = rep.num)
  mb <- fit_mb(pu$Xlog,     rep.num = rep.num)
  wg <- fit_wgcna(pu$Xlog)

  mods_gl <- greedy_modules(gl$A)
  mods_mb <- greedy_modules(mb$A)

  .pn  <- if (exists("KEEP_GLOBAL")) length(KEEP_GLOBAL[[kind]]) else NA_integer_
  m_gl <- metrics_from_adj(gl$A, mods_gl,   pool_n = .pn)
  m_mb <- metrics_from_adj(mb$A, mods_mb,   pool_n = .pn)
  m_wg <- metrics_from_adj(wg$A, wg$modules, pool_n = .pn)

  ## [repo 15.09.2026] the metric-row builder now lives in 02_functions.R as
  ## metric_row() -- it used to be defined here AND in 99_utils.R, and the two
  ## copies had drifted. The full documentation of each column moved with it.
  metrics <- list(Glasso = metric_row(m_gl), MB = metric_row(m_mb), WGCNA = metric_row(m_wg))

  # --- Table 1: Topology ---
  topo <- data.frame(Metric = names(metrics$Glasso),
                     Glasso_primary = metrics$Glasso, MB_crosscheck = metrics$MB,
                     WGCNA_modules = metrics$WGCNA, check.names = FALSE)
  write.csv(topo, file.path(outdir, "table1_topology_summary.csv"), row.names = FALSE)

  # --- Table 2: module assignment per method (+ ARI matrix) ---
  all_comp <- colnames(pu$Xlog)
  Mg <- align_modules(all_comp, mods_gl)
  Mm <- align_modules(all_comp, mods_mb)
  Mw <- align_modules(all_comp, wg$modules)
  modtab <- data.frame(Compound = all_comp, Use_group = ug_lookup[all_comp],
                       Glasso_module = Mg[all_comp], MB_module = Mm[all_comp],
                       WGCNA_module = as.character(wg$modules[all_comp]), check.names = FALSE)
  write.csv(modtab, file.path(outdir, "table2_module_assignments.csv"), row.names = FALSE)

  ari_mat <- matrix(1, 3, 3, dimnames = list(c("Glasso","MB","WGCNA"), c("Glasso","MB","WGCNA")))
  ari_mat["Glasso","MB"]    <- ari_mat["MB","Glasso"]    <- ari(Mg, Mm)
  ari_mat["Glasso","WGCNA"] <- ari_mat["WGCNA","Glasso"] <- ari(Mg, Mw)
  ari_mat["MB","WGCNA"]     <- ari_mat["WGCNA","MB"]      <- ari(Mm, Mw)
  write.csv(round(ari_mat, 3), file.path(outdir, "table2b_ARI_matrix.csv"))

  # --- Table 3: Null-model test per method ---
  ## [vRN 05.09.2026] TABLE 3 CAPTION -- HOW TO READ AND REPORT IT.
  ## Row order is deliberate: CONFIG first, CHUNGLU second, ER LAST AND SUPPLEMENTARY.
  ## The two degree-preserving nulls are the headline test: they are the H0 that the
  ## modularity statistic is defined against (k_i*k_j/2m, empirical degrees), so they are
  ## the only ones that test the same hypothesis the statistic assumes. The ER rows are
  ## retained for two reasons only -- continuity with tables reported before 31.08.2026,
  ## and as the worked counter-example that the choice of null IS the result (8 of 17
  ## units change verdict; 5 flip sign). In the thesis: quote config in the main text,
  ## move the ER modularity rows to the supplementary material, and keep ER for the
  ## average-path-length row, where it is the conventional and correct reference.
  ## [vRN 27.08.2026] THREE nulls, not one. igraph's modularity() scores a partition
  ## against the configuration-model expectation k_i*k_j/2m (hard-coded, degrees taken
  ## from the data), so comparing that statistic to an Erdos-Renyi ensemble tests a
  ## DIFFERENT H0 than the statistic itself assumes. `config` and `chunglu` are the
  ## internally consistent nulls; `er` is kept for continuity with earlier tables.
  ## Rank networks by `delta` (= observed - null_mean), NEVER by z: z is inflated
  ## wherever the null SD happens to be small (see the ER path-length column).
  ## [vRN 05.09.2026] ER PROVENANCE (meeting TODO 04.09.2026: where is the ER null used,
  ## why, with which ensemble, is there an original reference?).
  ##   WHERE   : only here, inside null_test_multi(nulls = c("config","chunglu","er")).
  ##             The legacy ER-only null_test() above is no longer called anywhere. ER
  ##             appears in no other part of the pipeline -- not in network construction,
  ##             edge selection, TU / risk work, RNA-seq or microbial branches.
  ##   ENSEMBLE: G(n, p) with n = vcount and p = observed edge_density; edge weights
  ##             resampled with replacement from the observed weight vector; fast-greedy
  ##             re-optimised on every draw; N draws -> z and empirical p.
  ##   REFERENCE: there is NO source reference for an external null test of Q. Hernandez
  ##             et al. (2021, ISME J 15:1722), the SGH paper this pipeline follows,
  ##             reports Q descriptively and correlates it with the stress ranking by
  ##             Spearman -- no null model at all. ER as a reference graph in the
  ##             microbial-network literature (Faust & Raes 2012, Nat Rev Microbiol
  ##             10:538) is the small-world reference for AVERAGE PATH LENGTH, which is
  ##             the one row where it remains appropriate here. For Q it is imported
  ##             convention, and an inconsistent one (see the note above).
  ##   VERDICT : keep ER for continuity of pre-31.08.2026 tables and as the worked
  ##             counter-example; config is the headline null. If Table 3 is ever cut for
  ##             space, drop the ER rows for Modularity and keep them for path length.
  nulltab <- do.call(rbind, lapply(c("Glasso","MB","WGCNA"), function(md) {
    A    <- switch(md, Glasso = gl$A,     MB = mb$A,     WGCNA = wg$A)
    memb <- switch(md, Glasso = mods_gl,  MB = mods_mb,  WGCNA = wg$modules)
    ## @NULL  THE NULL TEST. Change the nulls= vector here --------------------------------------------
    cbind(Model = md, null_test_multi(A, memb, N = n_null,
                                      nulls = c("config", "chunglu", "er")))
  }))
  nulltab$delta <- nulltab$observed - nulltab$null_mean        ## [vRN 27.08.2026]
  ## [vRN 31.08.2026] p_emp is an empirical p over N draws and cannot go below
  ## 1/(N+1). At N = 200 that floor is 0.00498, so a table full of 0.005 means
  ## "p < 0.005", not "p = 0.005". Carried as a column so the censoring is visible.
  ## @CAVEAT  empirical p cannot go below 1/(N+1): 0.005 means p < 0.005 ------------------------------
  nulltab$p_emp_floor <- 1 / (n_null + 1)
  num <- c("observed","null_mean","null_sd","z","p_emp","p_emp_floor","delta")
  nulltab[num] <- round(nulltab[num], 4)
  write.csv(nulltab, file.path(outdir, "table3_nullmodel_sensitivity.csv"), row.names = FALSE)

  ## [vPI2] --- Table 4 + Fig 4: absolute-risk-driver overlay (built on the glasso network) ---
  rd <- risk_driver_table(unit$site_ids, gl$A)
  write.csv(rd, file.path(outdir, "table4_risk_drivers.csv"), row.names = FALSE)
  plot_hub_vs_risk(rd, file.path(outdir, "fig4_hub_vs_riskdriver.png"), unit$label)
  rd_names <- rd$Compound[rd$absolute_risk_driver %in% TRUE]

  # --- Figures ---
  As    <- list(Glasso = gl$A, MB = mb$A, WGCNA = wg$A)
  by_mod <- list(Glasso = node_cols_module(gl$A, mods_gl),
                 MB     = node_cols_module(mb$A, mods_mb),
                 WGCNA  = node_cols_module(wg$A, wg$modules))
  by_ug  <- list(Glasso = node_cols_ug(gl$A), MB = node_cols_ug(mb$A), WGCNA = node_cols_ug(wg$A))
  mods   <- list(Glasso = mods_gl, MB = mods_mb, WGCNA = wg$modules)

  plot_modules(As, by_mod, mods, metrics,
               file.path(outdir, "fig1_network_panel.png"), unit$label, kind_label, risk = rd_names)
  plot_usegroups(As, by_ug, metrics,
                 file.path(outdir, "fig1b_network_panel_usegroup.png"),
                 unit$label, kind_label, ug_present_in(As), risk = rd_names)
  plot_wgcna_diag(wg, file.path(outdir, "fig2_wgcna_diagnostics.png"), unit$label, kind_label)
  plot_wgcna_dashboard(wg, file.path(outdir, "fig3_wgcna_dashboard.png"), unit$label, kind_label)

  ## [repo 15.09.2026] persist the FITTED OBJECTS, not just tables and PNGs.
  ## This one artefact lets 99_utils refresh figures WITHOUT refitting, lets S2
  ## drop its dependency on the external vRN_followups.R cache, and lets S3 read
  ## basin fits back instead of refitting them.
  save_fitted(list(A_glasso = gl$A, A_mb = mb$A, A_wgcna = wg$A,
                   mods_glasso = mods_gl, mods_mb = mods_mb, mods_wgcna = wg$modules,
                   lambda_glasso = gl$lambda, lambda_mb = mb$lambda,
                   stab_glasso = gl$stab, beta = wg$beta,
                   kind = kind, unit = unit, n_sites = pu$n_sites, n_comp = pu$n_comp,
                   rep.num = rep.num),
              kind, unit$safe)

  list(kind = kind, unit = unit$safe, label = unit$label, scale = scale_tag, topo = topo,
       ari = ari_mat, null = nulltab, beta = wg$beta, metrics = metrics,
       n_risk_drivers = sum(rd$absolute_risk_driver %in% TRUE),
       n_comp = pu$n_comp, n_sites = pu$n_sites)
}
```


This chunk takes a while (StARS subsampling + 100 weighted null graphs per network).

```{r run-all-networks, results='hide'}
summaries <- list()

# country-level sections (groups)
for (kind in c("conc", "tu")) {
  for (u in units) {
    res <- tryCatch(run_unit(kind, u, scale_tag = "group"),
                    error = function(e) { message("error on ", kind, " ", u$safe, ": ", conditionMessage(e)); NULL })
    if (!is.null(res)) summaries[[paste(kind, u$safe, sep = "_")]] <- res
  }
}

## [vPI2] river-basin-specific layer (all sites per basin, n >= BASIN_MIN_N)
for (kind in c("conc", "tu")) {
  for (u in basin_units) {
    res <- tryCatch(run_unit(kind, u, scale_tag = "basin"),
                    error = function(e) { message("error on ", kind, " ", u$safe, ": ", conditionMessage(e)); NULL })
    if (!is.null(res)) summaries[[paste(kind, u$safe, sep = "_")]] <- res
  }
}

## [vRN] latitude/longitude river sections (Upper/Middle/Lower incl. Elbe)
for (kind in c("conc", "tu")) {
  for (u in section_units) {
    res <- tryCatch(run_unit(kind, u, scale_tag = "section"),
                    error = function(e) { message("error on ", kind, " ", u$safe, ": ", conditionMessage(e)); NULL })
    if (!is.null(res)) summaries[[paste(kind, u$safe, sep = "_")]] <- res
  }
}

saveRDS(summaries, file.path(outroot, "_all_summaries.rds"))
```


# Pull everything together

```{r analysis-pipeline}

data_label <- function(kind) if (kind == "conc") "Concentration" else "Toxic unit (TRIDENT)"

topo_all <- do.call(rbind, lapply(summaries, function(s) {
  do.call(rbind, lapply(c("Glasso","MB","WGCNA"), function(md) {
    m <- s$metrics[[md]]
    data.frame(Data = data_label(s$kind), Scale = s$scale, Unit = s$label, Sites = s$n_sites, Model = md,
               Nodes = m["Nodes"], Edges = m["Edges"], Density = m["Density"],
             Mean_clustering = m["Mean clustering (unweighted)"], Modularity_Q = m["Modularity Q"],
               Avg_path_length = m["Avg path length"], row.names = NULL)
  }))
}))

ari_all <- do.call(rbind, lapply(summaries, function(s)
  data.frame(Data = data_label(s$kind), Scale = s$scale, Unit = s$label,
             ARI_Glasso_MB = s$ari["Glasso","MB"],
             ARI_Glasso_WGCNA = s$ari["Glasso","WGCNA"],
             ARI_MB_WGCNA = s$ari["MB","WGCNA"], row.names = NULL)))

null_all <- do.call(rbind, lapply(summaries, function(s)
  cbind(Data = data_label(s$kind), Scale = s$scale, Unit = s$label, s$null, row.names = NULL)))

## [vPI2] master risk-driver table across all units (concentration scale)
rd_all <- do.call(rbind, lapply(summaries, function(s) {
  if (s$kind != "conc") return(NULL)
  f <- file.path(outroot, "concentration", s$unit, "table4_risk_drivers.csv")
  if (!file.exists(f)) return(NULL)
  d <- read.csv(f, check.names = FALSE)
  d <- d[d$absolute_risk_driver %in% c(TRUE, "TRUE"), , drop = FALSE]
  if (!nrow(d)) return(NULL)
  cbind(Scale = s$scale, Unit = s$label, d)
}))

write.csv(topo_all, file.path(outroot, "master_topology_summary.csv"), row.names = FALSE)
write.csv(ari_all,  file.path(outroot, "master_ARI.csv"), row.names = FALSE)
write.csv(null_all, file.path(outroot, "master_nullmodel_sensitivity.csv"), row.names = FALSE)
if (!is.null(rd_all)) write.csv(rd_all, file.path(outroot, "master_risk_drivers.csv"), row.names = FALSE)

wb <- createWorkbook()
head_style <- createStyle(textDecoration = "bold", fgFill = "#264653", fontColour = "white", halign = "center")
add_sheet <- function(name, df) {
  addWorksheet(wb, name); writeData(wb, name, df, headerStyle = head_style)
  freezePane(wb, name, firstRow = TRUE); setColWidths(wb, name, cols = 1:ncol(df), widths = "auto")
}
add_sheet("Topology", topo_all); add_sheet("ARI", ari_all); add_sheet("NullModel", null_all)
if (!is.null(rd_all)) add_sheet("RiskDrivers", rd_all)
saveWorkbook(wb, file.path(outroot, "network_review_results.xlsx"), overwrite = TRUE)

dir.create(file.path(outroot, "data_for_marvin_upload"), showWarnings = FALSE)
write.csv(data.frame(Compound = rownames(conc_mat), Use_group = ug_lookup[rownames(conc_mat)],
                     conc_mat, check.names = FALSE),
          file.path(outroot, "data_for_marvin_upload", "raw_concentration_matrix_ngL.csv"), row.names = FALSE)
write.csv(data.frame(Compound = rownames(tu_mat), Use_group = ug_lookup[rownames(tu_mat)],
                     tu_mat, check.names = FALSE),
          file.path(outroot, "data_for_marvin_upload", "raw_toxicunit_matrix_TRIDENT.csv"), row.names = FALSE)
write.csv(data.frame(Compound = names(ec10_ngL), EC10_ngL = as.numeric(ec10_ngL)),
          file.path(outroot, "data_for_marvin_upload", "TRIDENT_EC10_ngL.csv"), row.names = FALSE)
write.csv(sites_meta, file.path(outroot, "data_for_marvin_upload", "site_metadata.csv"), row.names = FALSE)

## [repo 15.09.2026] run manifest: parameters + package versions + git commit,
## so every number in this output tree is traceable to its exact configuration.
write_run_manifest(extra = list(
  pipeline = "03_mainline",
  units_run = vapply(summaries, function(s) paste(s$kind, s$unit, sep = ":"), character(1))))
```


# Check-In on the Data

```{r sanity-check-on-the-data}

topo_all %>%
  group_by(Model) %>%
  summarise(units = sum(!is.na(Modularity_Q)),
            Q_over_0.3 = sum(Modularity_Q > 0.3, na.rm = TRUE),
            median_Q = round(median(Modularity_Q, na.rm = TRUE), 2))

null_all %>%
  filter(metric == "Modularity", !is.na(z)) %>%
  group_by(nullModel) %>%                                     ## [vRN 27.08.2026]
  summarise(runs = n(), z_over_2 = sum(z > 2), p_under_0.05 = sum(p_emp < 0.05),
            median_delta = round(median(delta, na.rm = TRUE), 3))
```

# Session info

```{r session-info}
sessionInfo()
```

