---
title: "S1 -- Method justification and robustness"
subtitle: "Why the prevalence filter is 10 %, why the node set is global, glasso vs MB, and how much the MDL/2 floor matters."
author: "Robert Neugebauer (vRN)"
date: "14.09.2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

This companion defends the choices made in the mainline script. Nothing here feeds Tables 1-4.

```{r split-source, message=FALSE}
## Everything below assumes the shared setup/data/functions are loaded.
## Repo-relative: knitr sets the working directory to this .Rmd file's folder.
sp <- file.path("..", "R")
source(file.path(sp, "00_setup.R"))
source(file.path(sp, "01_data.R"))
source(file.path(sp, "02_functions.R"))
```


## [vPI2] Prevalence-threshold diagnostic {#prev-diagnostic}

Because 58–77 % of the filtered matrix can still be tied non-detects at a 10 %
threshold, co-*absence* (two compounds jointly below MDL at the same sites) can
create edges that look like co-*occurrence*. This diagnostic sweeps the threshold
and reports, per unit: how many compounds survive, how much of the matrix is still
zero, and — most diagnostically — **edge fragility**: the fraction of candidate
correlations that vanish when recomputed using only the sites where *both*
compounds are detected (co-presence). A threshold where fragility stops falling is
where co-absence artefacts are gone. Review this before fixing `PREV_THRESH`.

```{r prev-diagnostic, fig.width = 10, fig.height = 4}

## [vRN 03.09.2026] NOTE: this sweep is deliberately WITHIN-unit. It is the
## evidence for the VALUE of PREV_THRESH, not for where the filter is applied.
## With NODESET_SCOPE = "global" that value is applied once over REF_SITES and
## the survivors become the node pool for every unit -- see the nodeset-audit
## chunk below.
prev_diagnostic <- function(kind, site_ids, thresholds = c(0.10, 0.20, 0.30, 0.40, 0.50)) {
  mat   <- if (kind == "conc") conc_mat else tu_mat
  sites <- intersect(site_ids, colnames(mat))
  full  <- mat[, sites, drop = FALSE]
  do.call(rbind, lapply(thresholds, function(pv) {
    md  <- max(2L, ceiling(pv * ncol(full)))
    sub <- full[rowSums(full > 0) >= md, , drop = FALSE]
    if (nrow(sub) < 5) return(data.frame(prev = pv, n_comp = nrow(sub),
                                         pct_zero = NA, frac_fragile = NA))
    ## [vPI3] same non-detect substitution as prep_unit(): replace only true zeros with
    ## MDL/2, leave detected values untouched, so the diagnostic reflects what the model sees
    floor_vec <- if (kind == "conc") {
      mdl_half_ngL[rownames(sub)]
    } else {
      fb <- 0.5 * min(sub[sub > 0])
      v  <- mdl_half_tu[rownames(sub)]
      v[is.na(v)] <- fb
      v
    }
    is_zero     <- sub == 0
    floor_mat   <- matrix(floor_vec, nrow = nrow(sub), ncol = ncol(sub))
    sub_floored <- sub
    sub_floored[is_zero] <- floor_mat[is_zero]
    X    <- log10(t(sub_floored))
    Rall <- suppressWarnings(cor(X))
    # co-presence-only correlation for the strongest candidate pairs
    ut  <- which(upper.tri(Rall) & abs(Rall) >= 0.3, arr.ind = TRUE)
    frag <- NA
    if (nrow(ut)) {
      keep <- 0L
      for (k in seq_len(nrow(ut))) {
        i <- ut[k, 1]; j <- ut[k, 2]
        both <- (sub[i, ] > 0) & (sub[j, ] > 0)
        if (sum(both) >= 4) {
          rr <- suppressWarnings(cor(X[both, i], X[both, j]))
          if (!is.na(rr) && abs(rr) >= 0.3) keep <- keep + 1L
        }
      }
      frag <- 1 - keep / nrow(ut)     # fraction of strong edges that do NOT survive co-presence
    }
    data.frame(prev = pv, n_comp = nrow(sub),
               pct_zero = round(mean(sub == 0) * 100, 1),
               frac_fragile = round(frag, 3))
  }))
}

# Run on two representative units (a country group and a whole basin)
diag_units <- list(groups[[3]], if (length(basin_units)) basin_units[[1]] else groups[[1]])
prev_tab <- do.call(rbind, lapply(diag_units, function(u)
  cbind(Unit = u$label, prev_diagnostic("conc", u$site_ids))))
dir.create(outroot, recursive = TRUE, showWarnings = FALSE)
write.csv(prev_tab, file.path(outroot, "prevalence_diagnostic.csv"), row.names = FALSE)
print(prev_tab)
```


## [vRN 03.09.2026] Node-set audit {#nodeset-audit}

What the fixed node set changes, unit by unit. `bar_old` is the number of
detections a compound needed under the old within-unit filter; it moved with
unit size, which is why `n_old` was not monotone in the nesting (a section
could carry more compounds than the basin containing it).

```{r nodeset-audit}
audit_units   <- c(units, basin_units, section_units)   # reach units are built later (vrn-refresh-figures)
nodeset_audit <- do.call(rbind, lapply(audit_units, function(u) {
  do.call(rbind, lapply(c("conc", "tu"), function(kind) {
    pu_old <- prep_unit(kind, u$site_ids, scope = "unit")
    pu_new <- prep_unit(kind, u$site_ids, scope = "global")
    ndet   <- rowSums(pu_new$cs > 0)
    data.frame(unit     = u$safe,
               kind     = kind,
               n_sites  = pu_new$n_sites,
               bar_old  = max(2L, ceiling(PREV_THRESH * pu_new$n_sites)),
               n_old    = pu_old$n_comp,
               n_new    = pu_new$n_comp,
               added    = length(setdiff(rownames(pu_new$cs), rownames(pu_old$cs))),
               removed  = length(setdiff(rownames(pu_old$cs), rownames(pu_new$cs))),
               n_lowdet = sum(ndet < 5),          # global nodes with < 5 detections here
               pct_zero = round(mean(pu_new$cs == 0) * 100, 1))
  }))
}))
dir.create(outroot, recursive = TRUE, showWarnings = FALSE)
write.csv(nodeset_audit, file.path(outroot, "nodeset_audit.csv"), row.names = FALSE)
print(subset(nodeset_audit, kind == "conc"))
```


# Model comparison with netCompare() — the MB robustness check

```{r netcompare, results='hide'}
library(NetCoMi)

compare_pair <- function(A1, A2) {
  if (sum(A1 != 0) == 0 || sum(A2 != 0) == 0) return(c(ARI = NA, Jacc_eigen = NA, Jacc_hub = NA))
  nodes <- colnames(A1); A2 <- A2[nodes, nodes]
  net <- netConstruct(data = A1, data2 = A2, dataType = "condDependence",
                      sparsMethod = "none", verbose = 0)
  pr  <- netAnalyze(net, clustMethod = "cluster_fast_greedy", hubPar = "eigenvector", hubQuant = 0.9)
  sm  <- summary(netCompare(pr, permTest = FALSE, verbose = FALSE))
  jac <- suppressWarnings(as.numeric(sm$jaccmat[, "Jacc"])); names(jac) <- rownames(sm$jaccmat)
  c(ARI        = as.numeric(sm$rand["ARI", "wholeNet"]),
    Jacc_eigen = unname(jac["eigenvec. centr."]),
    Jacc_hub   = unname(jac["hub taxa"]))
}

netcomp <- list()
all_run_units <- c(units, basin_units, section_units)   ## [vPI2]+[vRN] include basin + river-section layers
for (kind in c("conc", "tu")) {
  for (u in all_run_units) {
    pu <- prep_unit(kind, u$site_ids); if (pu$n_comp < 5) next
    A_gl <- tryCatch(fit_glasso(pu$Xlog)$A,
             error = function(e) matrix(0, pu$n_comp, pu$n_comp,
                       dimnames = list(colnames(pu$Xlog), colnames(pu$Xlog))))
    A_mb <- fit_mb(pu$Xlog)$A
    A_wg <- fit_wgcna(pu$Xlog)$A
    for (pair in list(c("Glasso", "MB"), c("Glasso", "WGCNA"), c("MB", "WGCNA"))) {
      A1 <- switch(pair[1], Glasso = A_gl, MB = A_mb)
      A2 <- switch(pair[2], MB = A_mb, WGCNA = A_wg)
      v  <- tryCatch(compare_pair(A1, A2), error = function(e) c(ARI = NA, Jacc_eigen = NA, Jacc_hub = NA))
      netcomp[[length(netcomp) + 1]] <- data.frame(
        Data = if (kind == "conc") "Concentration" else "Toxic unit (TRIDENT)",
        Unit = u$label, Pair = paste(pair, collapse = " vs "),
        ARI_clusters      = round(v["ARI"], 3),
        Jaccard_eigenHubs = round(v["Jacc_eigen"], 3),
        Jaccard_hubset    = round(v["Jacc_hub"], 3), row.names = NULL)
    }
  }
}
netcomp <- do.call(rbind, netcomp)
write.csv(netcomp, file.path(outroot, "model_comparison_netCompare.csv"), row.names = FALSE)
```


# [vRN] MDL/2 sensitivity — how much does the non-detect floor matter?

Recomputes the primary glasso network for one representative unit under three non-detect
schemes — `MDL/2` (the vPI3 default), the full `MDL`, and `drop-zeros` (only compounds with
no non-detects in the unit) — and reports edges, density, modularity and the module ARI
against the MDL/2 network. Stable numbers here mean the MDL/2 assumption is not driving the
topology.

```{r vrn-mdl-sensitivity, results='hide'}
mdl_sensitivity <- function(unit, kind = "conc") {
  M     <- if (kind == "conc") conc_mat else tu_mat
  sites <- intersect(unit$site_ids, colnames(M))
  sub0  <- M[, sites, drop = FALSE]
  min_detect <- max(2L, ceiling(PREV_THRESH * ncol(sub0)))
  sub0  <- sub0[rowSums(sub0 > 0) >= min_detect, , drop = FALSE]
  if (nrow(sub0) < 5) return(NULL)
  comps <- rownames(sub0)

  floor_half <- mdl_half_ngL[comps]; floor_half[is.na(floor_half)] <- DEFAULT_PSEUDO_CONC
  floor_full <- mdl_ngL[comps];      floor_full[is.na(floor_full)] <- 2 * DEFAULT_PSEUDO_CONC
  build <- function(fl) {
    is0 <- sub0 == 0; fm <- matrix(fl, nrow(sub0), ncol(sub0)); s <- sub0; s[is0] <- fm[is0]
    fit_glasso(log10(t(s)), rep.num = 10)$A
  }
  As <- list(`MDL/2` = build(floor_half), `MDL_full` = build(floor_full))
  nozero <- rowSums(sub0 == 0) == 0
  As[["drop_zeros"]] <- if (sum(nozero) >= 5) fit_glasso(log10(t(sub0[nozero, , drop = FALSE])), rep.num = 10)$A else NULL

  base <- As[["MDL/2"]]; mb <- greedy_modules(base)
  cmp <- function(A) {
    if (is.null(A)) return(c(edges = NA, density = NA, modularity = NA, ARI_vs_MDLhalf = NA))
    mm <- greedy_modules(A); mt <- metrics_from_adj(A, mm)
    common <- intersect(names(mb), names(mm))
    c(edges = mt$edges, density = round(mt$density, 4), modularity = round(mt$modularity, 3),
      ARI_vs_MDLhalf = round(ari(mb[common], mm[common]), 3))
  }
  do.call(rbind, lapply(names(As), function(nm) data.frame(scheme = nm, t(cmp(As[[nm]])), row.names = NULL)))
}

sens_unit <- if (length(basin_units)) basin_units[[1]] else units[[1]]
mdl_sens  <- tryCatch(mdl_sensitivity(sens_unit, "conc"),
                      error = function(e) { message("[vRN] MDL sensitivity failed: ", conditionMessage(e)); NULL })
if (!is.null(mdl_sens)) {
  write.csv(mdl_sens, file.path(outroot, sprintf("mdl_sensitivity_%s.csv", sens_unit$safe)), row.names = FALSE)
  print(mdl_sens)
}
```


## [vRN 05.09.2026] Optional: objective chemical families via ClassyFire {#classyfire}

`ug_lookup` above carries the PANGAEA **use group** (herbicide, pharmaceutical, ...), which
is a usage category, not a chemical one. A structural taxonomy is a genuinely independent
axis: it lets the module interpretation be tested against chemical family as well as
against use group and mode of action (A30), and it is assigned by an algorithm rather than
by hand. ClassyFire (Djoumbou Feunang et al. 2016, J Cheminform 8:61) provides exactly that
-- kingdom / superclass / class / subclass, derived from structure. `compound_inchikey`,
read from the PANGAEA sheet just above, is the lookup key.

This chunk is **`eval = FALSE` by design**: it is a network call over ~610 compounds, it is
slow and rate-limited, and it should be run once and cached. Nothing downstream depends on
it -- `cf_lookup` is an optional extra annotation column.

```{r vrn-classyfire, eval = FALSE}
## Batch ClassyFire classification, keyed on InChIKey, cached to disk.
## Run once; the CSV is the artefact and the chunk stays eval = FALSE afterwards.
## NB both endpoints below are third-party services that have changed and gone down
## before (the Wishart server especially). Check the cache before relying on it.
cf_cache <- file.path(outroot, "classyfire_lookup.csv")

classyfire_one <- function(ik) {
  ## preferred: the classyfireR package (CRAN), which wraps the ClassyFire REST API
  if (requireNamespace("classyfireR", quietly = TRUE)) {
    r <- try(classyfireR::get_classification(ik), silent = TRUE)
    if (!inherits(r, "try-error") && !is.null(r)) {
      cl <- try(classyfireR::classification(r), silent = TRUE)
      if (!inherits(cl, "try-error") && is.data.frame(cl) && nrow(cl))
        return(setNames(as.character(cl$Classification), tolower(as.character(cl$Level))))
    }
  }
  ## fallback: the Fiehn lab batch mirror, which serves pre-computed entities by InChIKey
  if (requireNamespace("jsonlite", quietly = TRUE)) {
    u <- sprintf("https://cfb.fiehnlab.ucdavis.edu/entities/%s.json", ik)
    j <- try(jsonlite::fromJSON(u), silent = TRUE)
    if (!inherits(j, "try-error") && is.list(j))
      return(c(kingdom    = j$kingdom$name,   superclass = j$superclass$name,
               class      = j$class$name,     subclass   = j$subclass$name))
  }
  NULL
}

if (file.exists(cf_cache)) {
  cf_tab <- read.csv(cf_cache, stringsAsFactors = FALSE)
} else {
  iks  <- unique(compound_inchikey[nzchar(compound_inchikey) & !is.na(compound_inchikey)])
  rows <- vector("list", length(iks))
  for (i in seq_along(iks)) {
    x <- classyfire_one(iks[i])
    pick <- function(nm) if (!is.null(x) && nm %in% names(x)) unname(x[[nm]]) else NA_character_
    rows[[i]] <- data.frame(InChIKey = iks[i], kingdom = pick("kingdom"),
                            superclass = pick("superclass"), class = pick("class"),
                            subclass = pick("subclass"), stringsAsFactors = FALSE)
    Sys.sleep(0.5)                       # be polite: the API is rate-limited
    if (i %% 25 == 0) message("[classyfire] ", i, "/", length(iks))
  }
  cf_tab <- do.call(rbind, rows)
  write.csv(cf_tab, cf_cache, row.names = FALSE)
}

## compound -> chemical class: the structural parallel of ug_lookup
cf_lookup <- setNames(cf_tab$class[match(compound_inchikey, cf_tab$InChIKey)], compounds)
message(sprintf("[classyfire] %d of %d compounds classified, %d distinct classes",
                sum(!is.na(cf_lookup)), length(cf_lookup),
                length(unique(stats::na.omit(cf_lookup)))))
```


# [repo] Automated sensitivity sweeps {#sensitivity-sweeps}

Added 15.09.2026. The setup comments record two open questions that were blocked
on manual re-runs: (i) A44 recommends `PREV_THRESH = 0.05`, but flipping the
default first needs the unit inventory re-run at 0.05 (only Basin_Elbe had been
tested, and the small units are where StARS collapsed in A37); (ii) the StARS
depth sweep (STARS_REPNUM = 100) covered only the three whole basins, not the
small section/reach units. Both are automated here so that changing a default is
a one-line edit with the evidence attached.

## Unit inventory across candidate prevalence thresholds

For each candidate `PREV_THRESH`: the global node pool (what `global_nodeset()`
admits), and per unit the number of pool compounds that survive the within-unit
`MIN_DETECT_UNIT` guard -- i.e. how large each unit's network would actually be,
and whether the small units stay viable (>= 5 compounds). Cheap: no fitting.

```{r vrn-unit-inventory}
unit_inventory <- function(thresholds = c(0.05, 0.10, 0.15, 0.20), kind = "conc") {
  all_units <- c(units, basin_units, section_units, reach_units)
  mat <- if (kind == "conc") conc_mat else tu_mat
  do.call(rbind, lapply(thresholds, function(pv) {
    pool <- global_nodeset(kind, prev = pv)
    do.call(rbind, lapply(all_units, function(u) {
      sites <- intersect(u$site_ids, colnames(mat))
      sub   <- mat[rownames(mat) %in% pool, sites, drop = FALSE]
      n_ok  <- sum(rowSums(sub > 0) >= MIN_DETECT_UNIT)
      data.frame(prev = pv, unit = u$safe, n_sites = length(sites),
                 pool_size = length(pool), n_comp_fit = n_ok,
                 viable = n_ok >= 5, row.names = NULL)
    }))
  }))
}

inv <- unit_inventory()
write.csv(inv, file.path(outroot, "unit_inventory_prevsweep.csv"), row.names = FALSE)
print(subset(inv, !viable))          # the units a threshold change would kill
print(subset(inv, unit == "Basin_Elbe"))
```

## StARS depth sweep for the SMALL units

The `STARS_REPNUM = 100` decision rests on a basin-scale sweep; the comment in
`00_setup.R` says the small units "were NOT in this sweep". This closes that gap
for the two smallest section units: lambda and edge count at
rep.num x seed. Instability shows up as lambda/edges varying within a rep.num.

```{r vrn-stars-depth-small-units, results='hide'}
stars_depth_units <- function(units_to_test, repnums = c(20, 50, 100),
                              seeds = c(42, 7, 13), kind = "conc") {
  do.call(rbind, lapply(units_to_test, function(u) {
    pu <- prep_unit(kind, u$site_ids)
    if (pu$n_comp < 5) return(NULL)
    do.call(rbind, lapply(repnums, function(rn) {
      do.call(rbind, lapply(seeds, function(sd) {
        gl <- fit_glasso(pu$Xlog, seed = sd, rep.num = rn)
        data.frame(unit = u$safe, n_sites = pu$n_sites, rep.num = rn, seed = sd,
                   lambda = gl$lambda, edges = sum(gl$A != 0) / 2,
                   row.names = NULL)
      }))
    }))
  }))
}

small2 <- section_units[order(sapply(section_units, function(u) length(u$site_ids)))][1:2]
stars_small <- tryCatch(stars_depth_units(small2),
                        error = function(e) { message("[vRN] StARS depth sweep failed: ",
                                                      conditionMessage(e)); NULL })
if (!is.null(stars_small)) {
  write.csv(stars_small, file.path(outroot, "stars_depth_small_units.csv"), row.names = FALSE)
  print(stars_small)
}
```

## Signed-edge check: does dropping the sign of partial correlations matter?

`abs(A)` discards the sign of the (few) negative edges -- 55 of 9452 (0.6 %)
when checked on 27.08.2026. `signed_edge_check()` (02_functions.R) turns "we
checked" into a per-unit number: the negative-edge share, and the ARI between
the module partition on |A| and on the positive-edges-only network. ARI ~ 1
means no module claim can depend on the dropped signs. Reads the cached mainline
fits; refits nothing.

```{r vrn-signed-edge-check}
signed_tab <- do.call(rbind, lapply(c(units, basin_units, section_units), function(u) {
  fc <- tryCatch(load_fitted("conc", u$safe), error = function(e) NULL)
  if (is.null(fc)) return(NULL)
  cbind(unit = u$safe, signed_edge_check(fc$A_glasso))
}))
if (!is.null(signed_tab)) {
  write.csv(signed_tab, file.path(outroot, "signed_edge_check.csv"), row.names = FALSE)
  print(signed_tab)
}
```
