---
title: "S3 -- Cross-basin comparison"
subtitle: "Rarefaction to a common sample size, module sharing, and the Elbe / Danube / Rhine panel."
author: "Robert Neugebauer (vRN)"
date: "14.09.2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r split-source, message=FALSE}
## Everything below assumes the shared setup/data/functions are loaded.
## Repo-relative: knitr sets the working directory to this .Rmd file's folder.
sp <- file.path("..", "R")
source(file.path(sp, "00_setup.R"))
source(file.path(sp, "01_data.R"))
source(file.path(sp, "02_functions.R"))
```


# [vRN] Cross-basin comparison at a common sample size (rarefaction)

Basins differ in n (~20–150 sites) and sparser networks are mechanically more modular,
so raw per-basin topology is not comparable. Here every basin is subsampled to a common
site count, glasso is refit `R` times, and topology is averaged. A pairwise
**module-sharing** index then asks whether the *same compounds co-cluster* across basins
(correlation of the two basins' averaged co-membership matrices on their shared compounds).

```{r vrn-crossbasin, results='hide'}
crossbasin_rarefied <- function(basins = basin_units, kind = "conc",
                                target_n = NULL, R = 25, seed = 42) {
  set.seed(seed)
  M  <- if (kind == "conc") conc_mat else tu_mat
  ns <- sapply(basins, function(b) length(intersect(b$site_ids, colnames(M))))
  if (is.null(target_n)) target_n <- max(BASIN_MIN_N, min(ns))
  message("[vRN] cross-basin rarefaction to n = ", target_n, " sites, R = ", R)

  rows <- list(); comemb <- list()
  for (b in basins) {
    sid_all <- intersect(b$site_ids, colnames(M))
    if (length(sid_all) < target_n) next
    comps <- rownames(prep_unit(kind, sid_all)$cs)                 # 11.08.2026 - fixed accumulator set
    acc <- matrix(0, length(comps), length(comps), dimnames = list(comps, comps)); cnt <- acc
    mets <- matrix(NA_real_, R, 4, dimnames = list(NULL, c("edges","density","modularity","mean_clust")))
    ## [vRN-fix 12.08.2026] draw every subsample BEFORE any fitting, so the draws are
    ## independent of whatever the fitter does to the RNG.
    sid_draws <- if (length(sid_all) <= target_n) rep(list(sid_all), R) else
                 lapply(seq_len(R), function(i) sample(sid_all, target_n))
    for (r in seq_len(R)) {
      sid <- sid_draws[[r]]
      pu  <- tryCatch(prep_unit(kind, sid), error = function(e) NULL); if (is.null(pu) || pu$n_comp < 5) next
      ## [vRN 05.09.2026] rep.num deliberately 10 here: this loop refits many times and both
      ## arms of the comparison use the same setting, so the RELATIVE result is unaffected.
      ## Do not quote absolute edge counts out of this chunk -- the Rhine sweep shows a
      ## single-seed lambda flip at 20, so 10 is below the stability floor for small units.
      A   <- tryCatch(fit_glasso(pu$Xlog, rep.num = 10)$A, error = function(e) NULL); if (is.null(A)) next
      mm  <- greedy_modules(A); mt <- metrics_from_adj(A, mm)
      mets[r, ] <- c(mt$edges, mt$density, mt$modularity, mt$mean_clust)
      nm <- intersect(names(mm), comps)
      if (length(nm) >= 2) { same <- outer(mm[nm], mm[nm], `==`);
        acc[nm, nm] <- acc[nm, nm] + same; cnt[nm, nm] <- cnt[nm, nm] + 1 }
    }
    comemb[[b$basin]] <- acc / pmax(cnt, 1)
    rows[[b$basin]] <- data.frame(
      Basin = b$basin, n_rarefied = target_n,
      edges         = round(mean(mets[,"edges"],      na.rm = TRUE), 1),
      density       = round(mean(mets[,"density"],     na.rm = TRUE), 4),
      modularity    = round(mean(mets[,"modularity"],  na.rm = TRUE), 3),
      modularity_sd = round(sd(  mets[,"modularity"],  na.rm = TRUE), 3),
      mean_clust    = round(mean(mets[,"mean_clust"],  na.rm = TRUE), 3), row.names = NULL)
  }
  tab <- if (length(rows)) do.call(rbind, rows) else NULL

  bn <- names(comemb); pair <- list()
  if (length(bn) >= 2) for (i in 1:(length(bn) - 1)) for (j in (i + 1):length(bn)) {
    sc <- intersect(rownames(comemb[[bn[i]]]), rownames(comemb[[bn[j]]]))
    if (length(sc) >= 5) {
      Xi <- comemb[[bn[i]]][sc, sc]; Xj <- comemb[[bn[j]]][sc, sc]; ut <- upper.tri(Xi)
      pair[[length(pair) + 1]] <- data.frame(
        Basin_A = bn[i], Basin_B = bn[j], n_shared_compounds = length(sc),
        module_sharing_r = round(suppressWarnings(cor(Xi[ut], Xj[ut])), 3), row.names = NULL)
    }
  }
  list(topology = tab, sharing = if (length(pair)) do.call(rbind, pair) else NULL)
}

cb <- tryCatch(crossbasin_rarefied(basin_units, "conc", R = 25),
               error = function(e) { message("[vRN] cross-basin failed: ", conditionMessage(e)); NULL })
if (!is.null(cb) && !is.null(cb$topology)) {
  write.csv(cb$topology, file.path(outroot, "crossbasin_rarefied_topology.csv"), row.names = FALSE)
  if (!is.null(cb$sharing)) write.csv(cb$sharing, file.path(outroot, "crossbasin_module_sharing.csv"), row.names = FALSE)
  print(cb$topology); print(cb$sharing)
  png(file.path(outroot, "crossbasin_rarefied_modularity.png"), width = 1500, height = 1000, res = 150)
  par(mar = c(7, 4.5, 4, 1))
  ymax <- max(cb$topology$modularity + cb$topology$modularity_sd, na.rm = TRUE) * 1.15
  bp <- barplot(cb$topology$modularity, names.arg = cb$topology$Basin, las = 2,
                ylim = c(0, ymax), col = "#457B9D", border = "grey30",
                ylab = "Modularity Q (glasso)",
                main = sprintf("Cross-basin modularity at common n = %d (rarefied, mean +/- sd)",
                               cb$topology$n_rarefied[1]))
  arrows(bp, cb$topology$modularity - cb$topology$modularity_sd,
         bp, cb$topology$modularity + cb$topology$modularity_sd,
         angle = 90, code = 3, length = 0.05, col = "grey20")
  dev.off()
}
```


# [vRN] Basin comparison — Elbe vs Danube vs Rhine on one glasso panel

The panels above compare *estimators* (glasso / MB / WGCNA) within one unit. This section
turns the comparison around: **one estimator (glasso, the primary), three basins side by
side**, so the figure reads as "same method, different rivers". Nodes are coloured by
compound use group and sized by eigenvector centrality exactly as in `fig1b`, so the two
figures can sit next to each other in the deck.

Networks used to be refit here because the per-unit run only wrote tables and PNGs.
[repo 15.09.2026] The mainline now saves `fitted_<kind>_<unit>.rds` per unit, so the
basin fits are READ BACK from `outroot` (same numbers as the mainline, zero refit cost);
only basins without a cached fit are refit. The fits are also cached in-session in
`basin_cmp`, so re-knitting costs nothing. Set `BASIN_CMP_REFIT <- TRUE` to force a refit.

```{r vrn-basin-glasso-fits, results='hide'}
## ---------------------------------------------------------------------------
## [vRN 14.08.2026] Cross-basin panel, part 1: fit the three basin networks once
## and cache them. Everything downstream (panel figure + bar plot) reads `basin_cmp`.
## ---------------------------------------------------------------------------

CMP_BASINS <- c("Elbe", "Danube", "Rhine")   # order = panel order, left to right
CMP_KIND   <- "conc"                          # "conc" (ng/L) or "tu" (TRIDENT EC10)
CMP_REPNUM <- STARS_REPNUM                    # [vRN 05.09.2026] was 20; see the StARS-depth block
if (!exists("BASIN_CMP_REFIT")) BASIN_CMP_REFIT <- FALSE

## site IDs per basin: reuse `basin_units` when this chunk runs inside the full workflow,
## otherwise fall back to sites_meta so the chunk also works in a half-loaded session.
cmp_site_ids <- function(bn) {
  if (exists("basin_units")) {
    hit <- Filter(function(u) identical(u$basin, bn), basin_units)
    if (length(hit)) return(hit[[1]]$site_ids)
  }
  sites_meta$site_id[sites_meta$river_basin == bn]
}

fit_one_basin <- function(bn, kind = CMP_KIND, rep.num = CMP_REPNUM) {
  sid <- cmp_site_ids(bn)

  ## [repo 15.09.2026] reuse the mainline fit when it is on disk: same network,
  ## no refit. The safe name mirrors 01_data.R's basin_units construction.
  if (exists("load_fitted")) {
    fc <- tryCatch(load_fitted(kind, paste0("Basin_", gsub("[^A-Za-z0-9]+", "_", bn))),
                   error = function(e) NULL)
    if (!is.null(fc)) {
      mt <- metrics_from_adj(fc$A_glasso, fc$mods_glasso)
      rd_names <- if (exists("risk_driver_table")) {
        rd <- tryCatch(risk_driver_table(sid, fc$A_glasso), error = function(e) NULL)
        if (is.null(rd)) character(0) else rd$Compound[rd$absolute_risk_driver %in% TRUE]
      } else character(0)
      message(sprintf("[vRN] %s: cached fit (n=%d sites, p=%d compounds, E=%d, Q=%.3f)",
                      bn, fc$n_sites, fc$n_comp, mt$edges, mt$modularity))
      return(list(basin = bn, A = fc$A_glasso, lambda = fc$lambda_glasso,
                  memb = fc$mods_glasso, metrics = mt,
                  n_sites = fc$n_sites, n_comp = fc$n_comp, risk = rd_names))
    }
  }

  pu  <- prep_unit(kind, sid)
  if (pu$n_comp < 5 || pu$n_sites < 4) { message("[vRN] skipping ", bn, " (too small)"); return(NULL) }

  gl   <- fit_glasso(pu$Xlog, rep.num = rep.num)
  memb <- greedy_modules(gl$A)
  mt   <- metrics_from_adj(gl$A, memb)

  ## absolute risk drivers, for the red-square overlay (only if the TU layer is loaded)
  rd_names <- if (exists("risk_driver_table")) {
    rd <- tryCatch(risk_driver_table(sid, gl$A), error = function(e) NULL)
    if (is.null(rd)) character(0) else rd$Compound[rd$absolute_risk_driver %in% TRUE]
  } else character(0)

  message(sprintf("[vRN] %s: n=%d sites, p=%d compounds, E=%d, Q=%.3f",
                  bn, pu$n_sites, pu$n_comp, mt$edges, mt$modularity))

  list(basin = bn, A = gl$A, lambda = gl$lambda, memb = memb, metrics = mt,
       n_sites = pu$n_sites, n_comp = pu$n_comp, risk = rd_names)
}

if (!exists("basin_cmp") || BASIN_CMP_REFIT) basin_cmp <- list()
for (bn in CMP_BASINS) {
  if (!is.null(basin_cmp[[bn]]) && !BASIN_CMP_REFIT) { message("[vRN] ", bn, " cached — skipping refit"); next }
  basin_cmp[[bn]] <- tryCatch(fit_one_basin(bn),
                              error = function(e) { message("[vRN] ", bn, " failed: ", conditionMessage(e)); NULL })
}
basin_cmp <- basin_cmp[!vapply(basin_cmp, is.null, logical(1))]
basin_cmp <- basin_cmp[intersect(CMP_BASINS, names(basin_cmp))]   # keep panel order

cmp_kind_label <- if (CMP_KIND == "conc") "Concentrations (ng/L)" else "Toxic units (TRIDENT EC10)"
cmp_dir <- file.path(outroot, "basin_comparison")
dir.create(cmp_dir, recursive = TRUE, showWarnings = FALSE)
```


```{r vrn-basin-glasso-panel, results='hide'}
## ---------------------------------------------------------------------------
## [vRN 14.08.2026] Cross-basin panel, part 2: the figure.
## Same layout/legend as plot_usegroups(), but the three panels are basins, not methods.
## ---------------------------------------------------------------------------

basin_panel_title <- function(b)
  sprintf("%s\nn=%d sites | p=%d compounds\nQ=%.2f | E=%d | dens=%.3f",
          b$basin, b$n_sites, b$n_comp,
          b$metrics$modularity, b$metrics$edges, b$metrics$density)

plot_basins_usegroups <- function(cmp, file, kind_label, seed = 42) {
  As <- lapply(cmp, `[[`, "A")
  ug_present <- ug_present_in(As)

  png(file, width = 800 * length(cmp), height = 1180, res = 150)
  on.exit(dev.off(), add = TRUE)
  layout(matrix(c(seq_along(cmp), rep(length(cmp) + 1, length(cmp))),
                nrow = 2, byrow = TRUE), heights = c(5.3, 1.7))
  par(mar = c(1.5, 1, 5.5, 1), oma = c(0, 0, 3, 0))

  ## the vRN draw_net() takes a risk= overlay; older copies of it do not, so check first
  has_risk <- "risk" %in% names(formals(draw_net))
  for (b in cmp) {
    ttl <- basin_panel_title(b)
    if (has_risk) draw_net(b$A, node_cols_ug(b$A), ttl, seed = seed,
                           risk = if (length(b$risk)) b$risk else NULL)
    else          draw_net(b$A, node_cols_ug(b$A), ttl, seed = seed)
  }

  par(mar = c(0.5, 1, 0.5, 1)); plot.new()
  ## horiz = TRUE silently clips when many use groups are present (pharmaceutical and
  ## pesticide fall off the left edge) -- wrap into two rows instead
  legend("top", ncol = ceiling(length(ug_present) / 2), bty = "n", cex = 1.0,
         title = "Node colour — compound use group",
         legend = ug_present, fill = ug_palette[ug_present], border = NA)
  legend("bottom", horiz = TRUE, bty = "n", cex = 1.05,
         legend = c("Node size = eigenvector centrality", "Hub = top 10% (black ring)",
                    "Violet ring = absolute risk driver", "Edge = association"),
         pch = c(21, 21, 21, NA), pt.bg = c("grey60", "grey60", "white", NA),
         pt.cex = c(1.6, 1.8, 1.8, NA), pt.lwd = c(1, 2.4, 3, NA),
         col = c("grey40", "black", RISK_RING, NA), lty = c(NA, NA, NA, 1))

  mtext(sprintf("Basin comparison — glasso (primary) — %s   (nodes coloured by use group)",
                kind_label), outer = TRUE, cex = 1.1, font = 2)
  invisible(file)
}

if (length(basin_cmp)) {
  f_panel <- file.path(cmp_dir, sprintf("fig5_basins_glasso_usegroup_%s.png", CMP_KIND))
  plot_basins_usegroups(basin_cmp, f_panel, cmp_kind_label)
  message("[vRN] basin panel written: ", f_panel)
}
```


## [vRN] Topology side by side

Modularity is **not** density-invariant: a sparser network is mechanically more modular, so
raw Q partly reports edge count rather than compartmentalisation. The second panel therefore
repeats Q after thinning every basin network down to the edge count of the sparsest one
(strongest |partial correlation| edges kept, modules recomputed on the thinned graph). Read
the density-matched bars, not the raw ones, when comparing basins — and note this is a
single fit per basin, not the rarefied resampling in the section above.

```{r vrn-basin-barplot, results='hide'}
## ---------------------------------------------------------------------------
## [vRN 14.08.2026] Cross-basin panel, part 3: topology bar plot
## Q (raw) | Q (density-matched) | compounds | edges
## ---------------------------------------------------------------------------

## thin a weighted adjacency matrix to exactly `target_e` strongest edges, then re-cluster
thin_to_edges <- function(A, target_e) {
  ut <- which(upper.tri(A) & A != 0, arr.ind = TRUE)
  if (!nrow(ut) || nrow(ut) <= target_e) return(A)
  w    <- abs(A[ut])
  keep <- ut[order(w, decreasing = TRUE)[seq_len(target_e)], , drop = FALSE]
  B <- matrix(0, nrow(A), ncol(A), dimnames = dimnames(A))
  B[keep] <- A[keep]
  (B + t(B))
}

basin_topology_table <- function(cmp) {
  e_min <- min(vapply(cmp, function(b) b$metrics$edges, numeric(1)))
  do.call(rbind, lapply(cmp, function(b) {
    Am <- thin_to_edges(b$A, e_min)
    mm <- greedy_modules(Am)
    Qm <- metrics_from_adj(Am, mm)$modularity
    data.frame(Basin = b$basin, n_sites = b$n_sites, n_compounds = b$n_comp,
               edges = b$metrics$edges, density = round(b$metrics$density, 4),
               modularity = round(b$metrics$modularity, 3),
               modularity_matched = round(Qm, 3), matched_at_edges = e_min,
               mean_clust = round(b$metrics$mean_clust, 3), row.names = NULL)
  }))
}

plot_basin_bars <- function(tab, file, kind_label) {
  cols <- setNames(c("#457B9D", "#2A9D8F", "#E9C46A", "#9B2226", "#F4A261")[seq_len(nrow(tab))], tab$Basin)

  panels <- list(
    list(v = tab$modularity,         lab = "Modularity Q",
         main = "(a) Modularity Q (as fitted)",           fmt = "%.2f"),
    list(v = tab$modularity_matched, lab = "Modularity Q, density-matched",
         main = sprintf("(b) Q at a common %.0f edges", tab$matched_at_edges[1]), fmt = "%.2f"),
    list(v = tab$n_compounds,        lab = "Compounds in network (nodes)",
         main = "(c) Network size",                       fmt = "%.0f"),
    list(v = tab$edges,              lab = "Edges",
         main = "(d) Edge count",                         fmt = "%.0f")
  )

  png(file, width = 2400, height = 800, res = 150)
  on.exit(dev.off(), add = TRUE)
  par(mfrow = c(1, length(panels)), mar = c(5.5, 4.8, 4, 1.2), oma = c(0, 0, 3.2, 0))
  for (p in panels) {
    ymax <- max(p$v, na.rm = TRUE) * 1.18
    bp <- barplot(p$v, names.arg = tab$Basin, las = 2, ylim = c(0, ymax),
                  col = cols[tab$Basin], border = "grey30",
                  ylab = p$lab, main = p$main, cex.names = 0.95)
    text(bp, p$v, labels = sprintf(p$fmt, p$v), pos = 3, cex = 0.95, col = "grey15", xpd = NA)
  }
  mtext(sprintf("Basin comparison — glasso topology — %s   (single fit per basin; n differs, see table)",
                kind_label), outer = TRUE, cex = 1.15, font = 2)
  invisible(file)
}

if (length(basin_cmp) >= 2) {
  basin_topo <- basin_topology_table(basin_cmp)
  print(basin_topo)
  write.csv(basin_topo, file.path(cmp_dir, sprintf("table5_basin_topology_%s.csv", CMP_KIND)),
            row.names = FALSE)
  f_bars <- file.path(cmp_dir, sprintf("fig6_basins_topology_bars_%s.png", CMP_KIND))
  plot_basin_bars(basin_topo, f_bars, cmp_kind_label)
  message("[vRN] basin bar plot written: ", f_bars)
}
```

