---
title: "S5 — The SGH test: modularity along the stress gradient"
subtitle: "Per-unit stress index from toxic units vs glasso modularity, across all spatial units; with module-stability bootstrap."
author: "Robert Neugebauer (vRN)"
date: "15.09.2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r split-source, message=FALSE}
## Everything below assumes the shared setup/data/functions are loaded.
## Repo-relative: knitr sets the working directory to this .Rmd file's folder.
sp <- file.path("..", "R")
source(file.path(sp, "00_setup.R"))
source(file.path(sp, "01_data.R"))
source(file.path(sp, "02_functions.R"))
```


# Rationale (read me first)

This is the test the thesis title promises. The Stress Gradient Hypothesis
(SGH), as formulated for co-occurrence networks by Hernandez et al. (2021,
ISME J 15:1722), predicts that **modular organisation declines along a stress
gradient**: modules insulate sub-communities from disturbance, and that
insulation is what increasing stress erodes. Their test is deliberately simple —
a per-site stress ranking against the modularity Q of each site's network, by
Spearman correlation. Here the unit of replication is the spatial unit
(country group, whole basin, river section, reach), the stress axis is the
mean per-site sum of toxic units (sum-TU = MEC/EC10, the mixture pressure), and
the response is the modularity of the unit's glasso network.

**The transfer caveat, stated openly.** What declines with stress in Hernandez
et al. is niche and interaction compartmentalisation between organisms. What
this pipeline measures is *source* compartmentalisation between chemicals — the
edge-level evidence (A30) says modules here are emission-source compartments,
not ecological niches. The SGH is therefore being tested by analogy: if stress
homogenises the chemical environment (fewer, dominant sources at high pressure;
a more even source mix at low pressure), source compartmentalisation should
decline with it. That is a hypothesis about emissions, not about interactions,
and the thesis should say so.

**The n-confound is controlled, not ignored.** Units differ in site count, and
sparser networks are mechanically more modular, so raw Q partly reports n
rather than stress. Two controls are built in: the regressions are weighted by
`n_sites` and include it as a covariate, and the rarefied cross-basin
comparison in S3 (every basin at a common n) provides the independent
cross-check. If the sign of the stress slope survives the n-adjustment, it is
not an artefact of unit size.


## Stress index per unit

```{r s5-stress-index}
## [repo 15.09.2026] S5 output folder. Everything this supplement writes lives
## under <outroot>/stress_gradient/, so the mainline output tree stays untouched.
out_s5 <- file.path(outroot, "stress_gradient")
dir.create(out_s5, recursive = TRUE, showWarnings = FALSE)

s5_all_units <- c(units, basin_units, section_units, reach_units)

## the stress axis, per unit: per-site sum-TU (mixture pressure), summarised as
## the mean, the 95th percentile (hotspot pressure), and the fraction of sites
## carrying at least one absolute risk driver (RQ >= RQ_THRESH, Inostroza 2024).
tu_ok <- nrow(tu_mat) > 0 && ncol(tu_mat) > 0
if (!tu_ok)
  message("[S5] tu_mat is empty (no EC10 source found) -- the stress axis cannot be built; downstream chunks will no-op.")

stress_tab <- if (tu_ok) do.call(rbind, lapply(s5_all_units, function(u) {
  sites <- intersect(u$site_ids, colnames(tu_mat))
  if (length(sites) < 4) return(NULL)
  sub_tu <- tu_mat[, sites, drop = FALSE]
  sum_tu <- colSums(sub_tu)
  data.frame(unit  = u$safe,
             label = u$label,
             basin = u$basin,
             n_sites_tu = length(sites),
             stress_mean = mean(sum_tu),
             stress_p95  = as.numeric(quantile(sum_tu, 0.95)),
             frac_sites_driver = mean(apply(sub_tu, 2, function(x) any(x >= RQ_THRESH))),
             row.names = NULL)
})) else NULL
if (!is.null(stress_tab)) print(stress_tab)
```


## The response: Q from the cached mainline fits

```{r s5-response}
## [repo 15.09.2026] reads the mainline cache only -- S5 fits nothing itself, so
## knit 03_mainline.Rmd first. Units without a cached fit are skipped, not refit.
scale_of <- function(safe) {
  if (grepl("^(Danube|Rhine)_", safe)) "group"
  else if (grepl("^Basin_", safe))     "basin"
  else if (grepl("^Sec_", safe))       "section"
  else if (grepl("^Reach_", safe))     "reach"
  else                                 "other"
}

resp_tab <- do.call(rbind, lapply(s5_all_units, function(u) {
  fc <- tryCatch(load_fitted("conc", u$safe), error = function(e) NULL)
  if (is.null(fc)) {
    message("[S5] no cached fit for ", u$safe, " -- skipping (knit 03_mainline.Rmd first)")
    return(NULL)
  }
  mt <- metrics_from_adj(fc$A_glasso, fc$mods_glasso)
  data.frame(unit    = u$safe,
             scale   = scale_of(u$safe),
             n_sites = fc$n_sites,
             n_comp  = fc$n_comp,
             edges   = mt$edges,
             density = round(mt$density, 4),
             Q       = round(mt$modularity, 4),
             row.names = NULL)
}))

## one row per unit: stress axis x network response, inner join on the safe name
sgh <- if (!is.null(stress_tab) && !is.null(resp_tab))
  merge(stress_tab, resp_tab, by = "unit") else NULL
if (!is.null(sgh)) print(sgh)
```


# The test

Three numbers carry the verdict: the **Spearman correlation** between stress
and Q (the direct analogue of the Hernandez et al. test — rank-based, so robust
to the skewed stress axis, but unweighted), a **weighted linear model** with
`n_sites` as weights (large units count more), and the same model with
**`n_sites` as a covariate** (does the stress slope survive once the mechanical
n-effect on Q is partialled out?). The unit table and the full model summaries
are written to `sgh_units.csv` / `sgh_models.txt` under the S5 folder.

```{r s5-test}
sgh_ok <- !is.null(sgh) && nrow(sgh) >= 5
if (sgh_ok) {
  rho_test <- suppressWarnings(cor.test(sgh$Q, sgh$stress_mean,
                                        method = "spearman", exact = FALSE))
  lm_w   <- lm(Q ~ stress_mean, data = sgh, weights = n_sites)
  lm_adj <- lm(Q ~ stress_mean + n_sites, data = sgh, weights = n_sites)

  print(rho_test)
  print(summary(lm_w))
  print(summary(lm_adj))

  write.csv(sgh, file.path(out_s5, "sgh_units.csv"), row.names = FALSE)
  writeLines(c("S5 -- the SGH test: modularity along the stress gradient",
               sprintf("units: %d   generated: %s", nrow(sgh), Sys.Date()),
               "",
               "== Spearman cor.test(Q, stress_mean) ==",
               capture.output(print(rho_test)),
               "",
               "== lm(Q ~ stress_mean, weights = n_sites) ==",
               capture.output(print(summary(lm_w))),
               "",
               "== lm(Q ~ stress_mean + n_sites, weights = n_sites) ==",
               capture.output(print(summary(lm_adj)))),
             file.path(out_s5, "sgh_models.txt"))
} else {
  message("[S5] fewer than 5 units with both stress and fit -- the test needs more rows; nothing written.")
}
```


## The figure

```{r s5-figure, results='hide'}
## [repo 15.09.2026] scatter of the SGH test: stress (x) vs Q (y), colour =
## spatial scale, size ~ sqrt(n_sites), labels = unit names without their
## scale prefix. Line = the n_sites-weighted fit. Base graphics, as everywhere
## in this repo.
if (sgh_ok) {
  scale_pal <- c(group = "#264653", basin = "#2A9D8F", section = "#E9C46A", reach = "#9B2226")
  p_adj <- coef(summary(lm_adj))["stress_mean", "Pr(>|t|)"]

  png(file.path(out_s5, "fig_sgh_scatter.png"), width = 1600, height = 1100, res = 150)
  par(mar = c(4.8, 4.8, 4.6, 1.2))
  plot(sgh$stress_mean, sgh$Q,
       pch = 21, bg = scale_pal[sgh$scale], col = "grey20",
       cex = 0.5 + sqrt(sgh$n_sites) / 5,
       xlab = "Stress index (mean sum-TU per site)",
       ylab = "Modularity Q (glasso, weighted)",
       main = "Does modularity decline as chemical stress rises?")
  abline(lm_w, col = "grey30", lty = 2, lwd = 2)
  text(sgh$stress_mean, sgh$Q,
       labels = sub("^([A-Za-z]+_)+", "", sgh$unit),
       pos = 3, cex = 0.6, col = "grey15", xpd = NA)
  legend("topright", bty = "n", cex = 0.9, title = "Spatial scale",
         legend = names(scale_pal), pch = 21, pt.bg = scale_pal, col = "grey20")
  mtext(sprintf("Spearman rho = %.2f (p = %.3g);  n-adjusted stress slope p = %.3g (weighted lm, line = unadjusted weighted fit)",
                unname(rho_test$estimate), rho_test$p.value, p_adj),
        side = 3, line = 0.3, cex = 0.8, col = "grey25")
  dev.off()
}
```


## Within-basin gradients: sections upstream to downstream

The cross-unit scatter mixes basins that differ in everything (size, land use,
compound set). The cleaner SGH contrast is *within* one basin: sections ordered
Upper to Lower form a natural gradient, and the two panels below trace stress
and Q along it, one line per basin.

```{r s5-section-panel, results='hide'}
## [repo 15.09.2026] sections ordered Upper -> Middle -> Lower (make_sections()
## in 01_data.R already orients "Upper" upstream per basin, splitting the Elbe
## and Rhine by latitude and the Danube by longitude). Only basins with >= 2
## sections can show a gradient at all.
pos_rank <- c(Upper = 1, Middle = 2, Lower = 3)
sec_by_basin <- split(section_units, vapply(section_units, `[[`, character(1), "basin"))
sec_by_basin <- Filter(function(x) length(x) >= 2, sec_by_basin)

if (sgh_ok && length(sec_by_basin)) {
  pd <- do.call(rbind, lapply(names(sec_by_basin), function(bn) {
    do.call(rbind, lapply(sec_by_basin[[bn]], function(u) {
      i <- match(u$safe, sgh$unit)
      if (is.na(i)) return(NULL)
      pos <- sub("^Sec_[A-Za-z0-9]+_", "", u$safe)
      data.frame(basin = bn, pos = pos, pos_i = unname(pos_rank[pos]),
                 stress = sgh$stress_mean[i], Q = sgh$Q[i], row.names = NULL)
    }))
  }))

  if (!is.null(pd) && nrow(pd)) {
    basin_cols <- c(Elbe = "#457B9D", Danube = "#2A9D8F", Rhine = "#E9C46A")
    col_of <- function(bn) if (bn %in% names(basin_cols)) basin_cols[[bn]] else "grey40"

    png(file.path(out_s5, "fig_sgh_within_basin.png"), width = 1800, height = 900, res = 150)
    par(mfrow = c(1, 2), mar = c(4.8, 4.8, 4, 1.2), oma = c(0, 0, 2.5, 0))
    for (pan in c("stress", "Q")) {
      yl <- range(pd[[pan]], na.rm = TRUE)
      if (diff(yl) == 0) yl <- yl + c(-0.5, 0.5)   # flat line still needs a finite window
      yl <- yl + c(-1, 1) * 0.08 * diff(yl)
      plot(NA, xlim = c(0.7, 3.3), ylim = yl, xaxt = "n",
           xlab = "River position (upstream to downstream)",
           ylab = if (pan == "stress") "Stress index (mean sum-TU per site)" else "Modularity Q (glasso)",
           main = if (pan == "stress") "(a) stress along the river" else "(b) modularity along the river")
      axis(1, at = 1:3, labels = c("Upper", "Middle", "Lower"))
      for (bn in unique(pd$basin)) {
        d <- pd[pd$basin == bn, ]
        d <- d[order(d$pos_i), ]
        lines(d$pos_i, d[[pan]], col = col_of(bn), lwd = 2)
        points(d$pos_i, d[[pan]], pch = 21, bg = col_of(bn), col = "grey20", cex = 1.6)
      }
      legend("topright", bty = "n", cex = 0.9, legend = unique(pd$basin),
             col = sapply(unique(pd$basin), col_of), pch = 19, lty = 1)
    }
    mtext("Within-basin gradient: river sections ordered upstream to downstream",
          outer = TRUE, cex = 1.05, font = 2)
    dev.off()
  }
}
```


## Module stability under site resampling

A gradient claim built on a partition that dissolves when 20 % of sites are
dropped would not survive review, so every basin partition gets a stability
number first: refit the glasso network on R = 25 site subsamples and report the
ARI of each subsample partition against the full-data partition.

```{r s5-module-stability, results='hide'}
## [repo 15.09.2026] SLOW: R = 25 glasso+StARS refits per basin on top of the
## full-data fit (~1-2 min per basin at rep.num = 20).
stab_tab <- do.call(rbind, lapply(basin_units, function(u) {
  ## safe = u$safe -> reference partition is the cached MAINLINE fit (StARS
  ## rep.num = 100), i.e. the partition the thesis reports, not a shallow refit
  bs <- tryCatch(bootstrap_module_stability("conc", u$site_ids, R = 25, rep.num = 20,
                                            safe = u$safe),
                 error = function(e) { message("[S5] stability bootstrap failed for ", u$safe,
                                               ": ", conditionMessage(e)); NULL })
  if (is.null(bs)) return(NULL)
  data.frame(unit = u$safe, mean_ari = round(bs$mean_ari, 3), sd_ari = round(bs$sd_ari, 3),
             R_done = bs$R_done, row.names = NULL)
}))
if (!is.null(stab_tab)) {
  write.csv(stab_tab, file.path(out_s5, "module_stability_bootstrap.csv"), row.names = FALSE)
  print(stab_tab)
}
```

How to read the table: an ARI near 1 means the same partition is recovered
under site resampling — the modules are a property of the data, not of which
sites happened to be sampled; values well below 1 mean the partition sits on
the degenerate plateau of near-optimal solutions (Good et al. 2010) and
module-level claims for that basin should be read accordingly.


# Session info

```{r session-info}
sessionInfo()
```
