---
title: "S6 — Which lens explains the modules: source vs structure vs mode of action"
subtitle: "Extends the A30 edge-level test to a third, structurally independent lens (ClassyFire), scored on the same networks and the same common node set."
author: "Robert Neugebauer (vRN)"
date: "15.09.2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r split-source, message=FALSE}
sp <- file.path("..", "R")
source(file.path(sp, "00_setup.R"))
source(file.path(sp, "01_data.R"))
source(file.path(sp, "02_functions.R"))
source(file.path(sp, "10_lenses_vRN.R"))
```

# Rationale (read me first)

`metrics_from_adj()`'s header records the A30 finding: the modules in these
networks are **emission-source compartments** — use group overwhelming, mode of
action adding nothing on top of it. A30 pitted two lenses against each other.
This supplement closes the question against a **third, structurally independent
lens** — ClassyFire chemical family — and scores every lens on the **same
networks and the same common node set**, so "most fitting" is a fair comparison
rather than an artefact of one lens labelling more compounds than another.

Each lens is scored three ways: the weighted **attribute modularity** of the
network when the lens is imposed as the partition (with a label-permutation
null, `Q_z`); nominal **assortativity**; and the A30 **edge-level nested
deviance**, generalised to report each lens's **marginal** deviance (alone) and
**unique** drop-one deviance (what it adds over the others). The headline is
**unique deviance**: a lens can look strong marginally only because it tracks
source, and the joint model removes that. A **negative coefficient** means
same-label pairs are *less* connected — significant but not a module generator.

**Prerequisites.** The structure lens needs the ClassyFire cache: run the
`{#classyfire}` chunk in `S1_method_justification.Rmd` once to write
`outputs_vRN/classyfire_lookup.csv` (set `CF_LEVEL <- "superclass"` in
`10_lenses_vRN.R` if class is too sparse). The MOA lens is optional — drop a
`raw_data/compound_moa.csv` (`compound`/`inchikey` + `moa`) and it is picked up
automatically; without it the comparison runs source-vs-structure.

# Run the comparison

```{r s6-run, message=TRUE}
## Spatial-gradient units: whole basins + along-river sections + reaches (conc).
lens_units <- c(basin_units, section_units, reach_units)

lenses <- assemble_lenses()
for (nm in names(lenses))
  message(sprintf("[S6] %-9s labels %d/%d compounds, %d groups", nm,
                  sum(!is.na(lenses[[nm]])), length(compounds),
                  length(unique(stats::na.omit(lenses[[nm]])))))

res <- lens_run(lens_units, kind = "conc", lenses = lenses)
```

```{r s6-verdict}
if (!is.null(res)) {
  print(res$verdict, row.names = FALSE)
} else {
  message("[S6] nothing scored — need >= 2 lenses and at least one fitted network. ",
          "Run the ClassyFire chunk in S1 and/or add raw_data/compound_moa.csv.")
}
```

```{r s6-figure, fig.width=11, fig.height=5}
if (!is.null(res)) {
  f <- lens_figure(res$master)
  message("[S6] figure written: ", f)
  knitr::include_graphics(f)
}
```

# How to read this

`master_lens_comparison.csv` has one row per unit × lens; `master_lens_verdict.csv`
collapses it to the winning lens per unit. A lens fits best when it carries the
**largest positive unique deviance** — it explains edges the others do not.
Three checks: a large **marginal** but near-zero **unique** deviance means the
lens only looked explanatory because it tracks source (the confound the joint
model removes); a **negative coefficient** (hatched bar in panel A) means
same-label pairs are less connected, so the lens is not a community generator;
and **coverage / `n_common`** guard against over-reading a lens the data can
barely label. If structure adds nothing over source here, the thesis sentence
strengthens: these modules encode emission source against **two** independent
alternatives (mechanism and chemical family), not one.
