---
title: "S8 -- Backbone sensitivity of the site-level tests"
subtitle: "Do the S7 results survive a different per-site backbone? LIONESS-on-glasso at the fixed aggregate StARS lambda vs the correlation-based estimators, on the same skeleton."
author: "Robert Neugebauer (vRN)"
date: "16.09.2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 13, fig.height = 8)
```

```{r split-source, message=FALSE}
## Everything below assumes the shared setup/data/functions are loaded AND that
## S6 has run TWICE: once normally (bonobo/lioness caches) and once with
## LIONESS_GLASSO <- TRUE (lioness_glasso caches). Repo-relative.
sp <- file.path("..", "R")
source(file.path(sp, "00_setup.R"))
source(file.path(sp, "01_data.R"))
source(file.path(sp, "02_functions.R"))
source(file.path(sp, "05_lioness.R"))
```


# Rationale (read me first)

S7's site-level tests all read per-site scores that were computed on ONE
backbone: correlation-based single-site networks (BONOBO / LIONESS on the
log-concentration matrix), summarised on the aggregate StARS skeleton. A
reviewer's first question is whether the headline -- a weak NEGATIVE
strength~stress coupling for BONOBO, a positive one for LIONESS -- is a
property of the data or of that backbone choice.

S8 repeats the key S7 tests with the third estimator S6 already provides:
**lioness_glasso**, the per-site partial-correlation companion to the mainline
network. It refits the glasso once per site at the FIXED aggregate StARS
lambda (leave-one-out), applies the LIONESS extrapolation
`e_q = n * A_agg - (n-1) * A_loo` to the partial correlations, and restricts
the result to the aggregate skeleton -- so it changes BOTH the association
measure (partial correlation instead of correlation) AND the extrapolation,
while keeping the skeleton fixed. BONOBO strengths are recomputed on the same
skeleton alongside, so the two sit on an identical footing.

Read S8 as a robustness bracket, not as a fourth method: if the S7 sign
pattern (BONOBO negative, LIONESS positive, UDF decides between them) were an
artifact of the correlation backbone, it should not reappear here.


## Parameters

```{r s8-params}
S8_NPERM   <- 5000L   # permutation draws for the null distributions
S8_MIN_UDF <- 10L     # minimum sites with UDF for a per-basin UDF test
S8_SEED    <- 42L
```


## Per-site analysis table (glasso backbone)

```{r s8-table}
## [S8] the lioness_glasso caches only exist if S6 was run with
## LIONESS_GLASSO <- TRUE -- fail loudly with instructions otherwise.
s6_units <- c(units, basin_units, section_units)
have_lg <- vapply(s6_units, function(u)
  !is.null(load_single_nets("conc", u$safe, "lioness_glasso")), logical(1))
if (!any(have_lg))
  stop("[S8] no lioness_glasso caches found. Re-knit S6_single_networks.Rmd ",
       "with LIONESS_GLASSO <- TRUE first (costs n glasso solves per unit), ",
       "then come back here.")

s8 <- do.call(rbind, lapply(s6_units[have_lg], function(u) {
  fit <- load_fitted("conc", u$safe)
  lg  <- load_single_nets("conc", u$safe, "lioness_glasso")
  bo  <- load_single_nets("conc", u$safe, "bonobo")
  if (is.null(fit) || is.null(lg) || is.null(bo)) return(NULL)
  s_lg <- single_site_summary(lg$nets, fit$A_glasso, fit$mods_glasso)
  s_bo <- single_site_summary(bo$nets, fit$A_glasso, fit$mods_glasso)
  data.frame(kind = "conc", unit = u$safe, label = u$label, site = s_lg$site,
             strength_lionessglasso = s_lg$strength_skeleton,
             strength_bonobo        = s_bo$strength_skeleton,
             stringsAsFactors = FALSE)
}))

## same stress axis and external covariates as S7
sum_tu <- if (ncol(tu_mat) > 0) colSums(tu_mat) else setNames(rep(NA_real_, ncol(conc_mat)), colnames(conc_mat))
s8$sum_tu <- sum_tu[s8$site]
udf_col <- "Urban discharge fraction (UDF)"
so_col  <- "Stream order by Strahler"
meta <- as.data.frame(sites_meta)
s8$UDF          <- as.numeric(meta[match(s8$site, meta$site_id), udf_col])
s8$stream_order <- as.numeric(meta[match(s8$site, meta$site_id), so_col])
s8$basin        <- meta[match(s8$site, meta$site_id), "river_basin"]

## trivial per-site scores, same definition as S7
pool <- KEEP_GLOBAL$conc
sub_pool <- conc_mat[pool, , drop = FALSE]
fl <- sub_pool
is_zero <- fl == 0
fl[is_zero] <- matrix(mdl_half_ngL[pool], nrow = length(pool), ncol = ncol(fl))[is_zero]
logf <- log10(fl)
s8$n_det    <- colSums(sub_pool > 0)[s8$site]
s8$mean_log <- colMeans(logf)[s8$site]

write.csv(s8, file.path(outroot, "tableS8_site_table.csv"), row.names = FALSE)
message("[S8] site table: ", nrow(s8), " site x unit rows over ",
        length(unique(s8$unit)), " units with lioness_glasso caches")
```


## Pooled inference and the permutation null (glasso backbone)

```{r s8-global}
## [S8] identical machinery to S7, only the strength column changes. The null
## shuffles sum-TU WITHIN each unit; the same null serves both estimators.
set.seed(S8_SEED)
units_u <- sort(unique(s8$unit))
fast_rho <- function(ry, x) cor(ry, rank(x, ties.method = "average"))
stouffer <- function(rhos, n) {
  z <- atanh(pmax(pmin(rhos, .999), -.999)) * sqrt(n - 3)
  sum(z) / sqrt(length(rhos))
}

ranks_lg <- lapply(units_u, function(u) rank(s8$strength_lionessglasso[s8$unit == u], ties.method = "average"))
ranks_bo <- lapply(units_u, function(u) rank(s8$strength_bonobo[s8$unit == u], ties.method = "average"))
x_vals   <- lapply(units_u, function(u) s8$sum_tu[s8$unit == u])
ns       <- vapply(x_vals, length, integer(1))

obs_lg <- mapply(fast_rho, ranks_lg, x_vals)
obs_bo <- mapply(fast_rho, ranks_bo, x_vals)
null_rhos <- replicate(S8_NPERM,
  mapply(function(ry, x) fast_rho(ry, sample(x)), ranks_lg, x_vals))
## [S8-fix 17.09.2026] same single-unit guard as S7: replicate() simplifies
## to a vector for one unit; force the matrix shape (no-op for full runs).
null_rhos <- matrix(null_rhos, nrow = length(ranks_lg))
p_emp_lg <- rowMeans(abs(null_rhos) >= abs(obs_lg))

obs_stats  <- c(neg = sum(obs_lg < 0), med = median(obs_lg), Z = stouffer(obs_lg, ns))
null_stats <- rbind(neg = colSums(null_rhos < 0),
                    med = apply(null_rhos, 2, median),
                    Z   = apply(null_rhos, 2, stouffer, n = ns))
p_neg <- mean(null_stats["neg", ] >= obs_stats["neg"])
p_med <- mean(null_stats["med", ] <= obs_stats["med"])
p_Z   <- mean(null_stats["Z", ]   <= obs_stats["Z"])
Z_bo  <- stouffer(obs_bo, ns)

per_unit <- data.frame(unit = units_u, n_sites = ns,
                       rho_lionessglasso = round(obs_lg, 3),
                       p_emp_lionessglasso = round(p_emp_lg, 4),
                       rho_bonobo_same_skeleton = round(obs_bo, 3),
                       row.names = NULL)
write.csv(per_unit, file.path(outroot, "tableS8_per_unit_permutation.csv"), row.names = FALSE)

global_tests <- data.frame(
  test = c("sign count negative (lioness_glasso)", "median rho (lioness_glasso)",
           "Stouffer Z (lioness_glasso)", "Stouffer Z (lioness_glasso)",
           "Stouffer Z (bonobo, same skeleton)"),
  statistic = round(c(obs_stats["neg"], obs_stats["med"], obs_stats["Z"], obs_stats["Z"], Z_bo), 4),
  p_value   = signif(c(p_neg, p_med, 2 * pnorm(-abs(obs_stats["Z"])), p_Z,
                       2 * pnorm(-abs(Z_bo))), 3),
  method    = c("permutation", "permutation", "normal theory", "permutation", "normal theory"),
  row.names = NULL)
write.csv(global_tests, file.path(outroot, "tableS8_global_tests.csv"), row.names = FALSE)
print(global_tests)
```

## External pressure proxy (UDF) and the biomarker test

```{r s8-external}
## [S8] basin units only, as in S7. The question that decided between BONOBO
## and LIONESS in S7 was the EXTERNAL proxy -- repeat it for the glasso
## backbone.
B <- s8[grepl("^Basin_", s8$unit), ]
B$z_strength <- ave(B$strength_lionessglasso, B$basin, FUN = function(s) as.numeric(scale(s)))
B$log_tu     <- log1p(B$sum_tu)
U <- B[!is.na(B$UDF), ]

udf_rows <- do.call(rbind, lapply(split(U, U$basin), function(d) {
  if (nrow(d) < S8_MIN_UDF) return(NULL)
  data.frame(basin = d$basin[1], n = nrow(d),
             rho_lionessglasso_strength_UDF = round(cor(d$strength_lionessglasso, d$UDF, method = "spearman"), 3),
             p_lionessglasso = signif(cor.test(d$strength_lionessglasso, d$UDF, method = "spearman")$p.value, 3),
             rho_bonobo_strength_UDF = round(cor(d$strength_bonobo, d$UDF, method = "spearman"), 3),
             p_bonobo = signif(cor.test(d$strength_bonobo, d$UDF, method = "spearman")$p.value, 3),
             row.names = NULL)
}))
write.csv(udf_rows, file.path(outroot, "tableS8_external_proxy_UDF.csv"), row.names = FALSE)
print(udf_rows)

fe_ols <- lm(z_strength ~ log_tu + basin, data = B)
fe_rlm <- MASS::rlm(z_strength ~ log_tu + basin, data = B)
t_rlm  <- summary(fe_rlm)$coefficients["log_tu", "t value"]
cat(sprintf("[S8] pooled slope log(1+sum-TU) [lioness_glasso]:  OLS %+.4f (p %.3f)   Huber %+.4f (p %.3f)\n",
            coef(fe_ols)["log_tu"], summary(fe_ols)$coefficients["log_tu", 4],
            coef(fe_rlm)["log_tu"], 2 * pnorm(-abs(t_rlm))))

Ub <- U[U$basin %in% udf_rows$basin[udf_rows$n >= 30], ]
Ub$z_udf <- ave(Ub$UDF, Ub$basin, FUN = function(s) as.numeric(scale(s)))
m0 <- lm(z_udf ~ log_tu + basin, data = Ub)
m1 <- lm(z_udf ~ log_tu + z_strength + basin, data = Ub)
biomarker <- anova(m0, m1)
cat(sprintf("[S8] biomarker test (UDF ~ log_tu vs + lioness_glasso strength): F = %.2f, p = %.3f\n",
            biomarker$F[2], biomarker$`Pr(>F)`[2]))
```


## Trivial-score benchmark

```{r s8-benchmark}
partial_spearman <- function(d, y, x, covs) {
  rres <- function(v) {
    X <- model.matrix(~ ., data = as.data.frame(lapply(d[covs], rank)))[, -1]
    residuals(lm(rank(v) ~ X))
  }
  cor(rres(d[[y]]), rres(d[[x]]), method = "spearman")
}
bench <- do.call(rbind, lapply(split(B, B$basin), function(d) {
  data.frame(basin = d$basin[1], n = nrow(d),
             raw_rho = round(cor(d$strength_lionessglasso, d$sum_tu, method = "spearman"), 3),
             partial_rho_given_trivial = round(
               partial_spearman(d, "strength_lionessglasso", "sum_tu", c("n_det", "mean_log")), 3),
             row.names = NULL)
}))
write.csv(bench, file.path(outroot, "tableS8_benchmark.csv"), row.names = FALSE)
print(bench)
```


## Side-by-side with S7

```{r s8-compare}
## [S8] put the three estimators next to each other. The S7 tables are read
## back from disk when present; otherwise only the S8 side is shown.
s7_global <- file.path(outroot, "tableS7_global_tests.csv")
if (file.exists(s7_global)) {
  s7g <- read.csv(s7_global, stringsAsFactors = FALSE)
  compare <- rbind(
    data.frame(backbone = "correlation (S7)", s7g[grepl("Stouffer|sign|median", s7g$test), ]),
    data.frame(backbone = "glasso (S8)", global_tests))
  print(compare)
}

png(file.path(outroot, "figS8_backbone_sensitivity.png"), width = 2200, height = 1100, res = 150)
par(mfrow = c(1, 2), mar = c(4.6, 10, 3.4, 1))

## (a) per-unit rho: glasso backbone vs correlation-backbone BONOBO/LIONESS
s7_unit <- file.path(outroot, "tableS7_per_unit_permutation.csv")
ord <- order(obs_lg)
plot(NA, xlim = range(c(obs_lg, obs_bo), na.rm = TRUE) * 1.1,
     ylim = c(0.5, length(units_u) + 0.5), yaxt = "n",
     xlab = "Spearman rho: per-site strength vs site sum-TU", ylab = "",
     main = "(a) per-unit rho: three estimators, one question")
axis(2, at = seq_along(units_u), labels = units_u[ord], las = 1, cex.axis = 0.7)
abline(v = 0, col = "grey60")
points(obs_lg[ord], seq_along(units_u) + 0.18, pch = 17, col = "#9B2226", cex = 1.2)
points(obs_bo[ord], seq_along(units_u) - 0.18, pch = 19, col = "#2A9D8F", cex = 1.2)
if (file.exists(s7_unit)) {
  s7u <- read.csv(s7_unit, stringsAsFactors = FALSE)
  rho_lio_s7 <- s7u$rho_lioness[match(units_u, s7u$unit)]
  points(rho_lio_s7[ord], seq_along(units_u), pch = 15, col = "#E9C46A", cex = 1.1)
}
legend("bottomright", bty = "n", cex = 0.85,
       legend = c("lioness_glasso (S8, glasso backbone)",
                  "BONOBO (same skeleton)",
                  "LIONESS correlation (S7)"),
       pch = c(17, 19, 15), col = c("#9B2226", "#2A9D8F", "#E9C46A"))

## (b) per-site strengths: glasso vs correlation backbone
par(mar = c(4.6, 4.8, 3.4, 1))
plot(s8$strength_bonobo, s8$strength_lionessglasso, pch = 19, cex = 0.5,
     col = adjustcolor("#264653", alpha.f = 0.35),
     xlab = "BONOBO strength (correlation backbone)",
     ylab = "lioness_glasso strength (glasso backbone)",
     main = sprintf("(b) per-site agreement (Spearman %.2f)",
                    cor(s8$strength_bonobo, s8$strength_lionessglasso, method = "spearman")))
abline(0, 1, lty = 3, col = "grey60")
dev.off()
```


## How to read the output

- `tableS8_per_unit_permutation.csv` — per-unit rho for the glasso-backbone
  estimator with its permutation p-value, plus BONOBO on the SAME skeleton.
- `tableS8_global_tests.csv` — the pooled evidence for lioness_glasso. Compare
  the sign and the permutation p-values against S7's `tableS7_global_tests.csv`.
- `tableS8_external_proxy_UDF.csv` — the circularity-free check on the glasso
  backbone: strength vs UDF per basin, lioness_glasso and BONOBO side by side.
- `tableS8_benchmark.csv` — raw vs partial rho given the trivial scores.
- `figS8_backbone_sensitivity.png` — the two panels above.

Interpretation guide. Three readings are possible, and the benchmark decides
between them:

1. lioness_glasso lands on the BONOBO side (negative pooled rho, negative UDF
   rho) -> the S7 conclusion is not a correlation-backbone artifact.
2. lioness_glasso lands on the LIONESS side (positive) AND its partial rho
   given the trivial scores (detection count, mean log-concentration) stays
   positive -> the sign is genuinely backbone-dependent; report the coupling
   as fragile to BOTH estimator and backbone.
3. lioness_glasso lands on the LIONESS side but its partial rho collapses
   toward zero -> the glasso-backbone strength is largely a detection-
   richness score (influential, detection-rich sites perturb the glasso fit
   most), and the correlation-backbone BONOBO result remains the only signal
   that survives trivial-score controls.

The delivered run (see results/s8 in the commit package) lands on reading 3:
lioness_glasso is strongly positive (Stouffer Z = +15.5, UDF rho = +0.39 in
the Elbe) but its per-basin partial rhos given the trivial scores collapse to
~0 or negative in 4 of 5 basins, and per-site agreement between the two
backbones is essentially zero (Spearman 0.09). BONOBO recomputed on the same
skeleton keeps its negative sign (Z = -1.66), attenuated versus S7's -2.56 --
so the magnitude of the BONOBO coupling is skeleton-sensitive even though its
direction is not.

Numerical provenance note: the delivered S8 numbers were produced with a
Python replication of this exact analysis (StARS with rep.num = 20, the repo's
own robustness convention, instead of 100; sklearn/ADMM glasso solvers instead
of huge -- 9 of 50 Rhine_Lower leave-one-out fits diverge under coordinate
descent because of near-perfectly correlated compound pairs at p > n, and are
solved with a validated ADMM fallback, 99.98% edge agreement). The R code in
this file reproduces the same analysis with huge throughout when knitted.


# Session info

```{r session-info}
sessionInfo()
```
