---
title: "S9 -- Paired chemical + microbial simulation with known ground truth"
subtitle: "Can the single-site network pipeline recover structural decoupling, and how large must a future same-site study be to detect a structure-driven biological response?"
author: "Robert Neugebauer (vRN)"
date: "17.09.2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 13, fig.height = 9)
```

```{r split-source, message=FALSE}
## Everything below assumes the shared setup/data/functions are loaded.
## S9 does NOT need S6 caches: it calibrates from prep_unit() output and,
## optionally, the mainline's cached Elbe glasso fit for the skeleton.
## Repo-relative: knitr sets the working directory to this .Rmd file's folder.
sp <- file.path("..", "R")
source(file.path(sp, "00_setup.R"))
source(file.path(sp, "01_data.R"))
source(file.path(sp, "02_functions.R"))
source(file.path(sp, "05_lioness.R"))
```


# Rationale (read me first)

S7/S8 established what the single-site networks do on the REAL data: BONOBO
strength on the correlation backbone carries a stress signal that survives
trivial-score controls, LIONESS is leverage-confounded, and the glasso backbone
tracks detection richness. What real data can never tell us is whether the
pipeline would recover a structure-driven biological response **if one existed**
— because in reality we do not know which sites are truly structurally
atypical, and we have no co-located biological measurements.

S9 therefore builds a synthetic world in which the truth is known:

1. **Calibrated chemistry.** Sites are drawn from a multivariate normal whose
   correlation matrix, variances, means, MDL floors and sum-TU distribution are
   all taken from the REAL Elbe `conc` unit (`prep_unit()` output). The only
   invented ingredient is a per-site **decoupling parameter kappa_q**: site q's
   covariance is the convex combination
   `Sigma_q = (1 - kappa_q) * Sigma + kappa_q * Sigma_broken`, where
   `Sigma_broken` has half of the real correlations shrunk toward zero. High
   kappa = the site's mixture does not follow the basin's co-occurrence rules.
   We KNOW which sites these are, because we made them. kappa increases with
   stress (plus noise), mimicking the SGH-motivated idea that stressed sites
   carry idiosyncratic mixtures. The simulated concentrations are then put
   through the SAME damage the real data suffer: censoring at the real MDLs,
   MDL/2 flooring, log10 transform.
2. **Calibrated microbiome.** A 300-taxon community is drawn from a
   Dirichlet-multinomial (the standard compositionality-aware count model).
   Taxa are pre-declared sensitive / tolerant / neutral (the SPEAR idea), and
   their expected abundances respond to the chemistry through two channels:
   beta1 * z(sum-TU) (the AMOUNT channel) and beta2 * z(kappa) (the STRUCTURE
   channel). Four scenarios: H0 (no response), H_amount (beta1 only),
   H_structure (beta2 only), H_both.
3. **Recovery.** The unmodified S6 estimators (`lioness_cor`, `bonobo_cor`)
   are run on the simulated chemistry. Does `strength_skeleton` track the TRUE
   kappa? Do the truly-decoupled sites rank at the bottom (ROC/AUC)? This tests
   our S7/S8 interpretation on known ground truth.
4. **Biomarker test with known truth.** The exact S7 model comparison
   (`metric ~ z(tu)` vs `metric ~ z(tu) + z(strength)`, F-test) is replayed per
   scenario. In H_structure strength MUST win; in H_amount it MUST NOT. If it
   misbehaves here, the real-data null result (F ~ 0) needs reinterpretation.
5. **Power grid.** Repeating (3)+(4) over n sites x beta2 gives the deliverable
   that designs the follow-up study: "a same-site chemical + eDNA study with
   n = X sites detects a structure effect of size Y with 80% power."

@CAVEAT  This is an IDENTIFIABILITY study, not evidence about reality. Every
distributional choice is copied from the Elbe data; the only invented things
are kappa and the betas — the levers under test. sum-TU is imposed from the
real distribution as the stress axis (computing TU from simulated
concentrations would require per-compound EC10s and adds nothing here).


## Parameters

```{r s9-params}
S9_SEED       <- 42L     # master seed; every sim draws from this stream
S9_NREP       <- 20L     # replicates per grid cell
S9_GRID_N     <- c(50L, 100L, 200L, 500L)   # sites per simulated study
S9_GRID_BETA2 <- c(0, 0.3, 0.6, 1.0)        # structure-channel effect sizes
S9_NTAXA      <- 300L    # simulated microbial taxa
S9_DEPTH      <- 10000L  # sequencing depth per site
S9_DM_PHI     <- 10      # Dirichlet-multinomial concentration (overdispersion)
S9_BETA_KAPPA <- 1.5     # how strongly decoupling rises with stress (logit slope)
S9_KAPPA_NOISE<- 1.0     # noise in the stress->kappa link (keeps kappa != f(TU))
S9_ALPHA_MEAN <- 0.5     # mean shift of log-concentrations along PC1 per stress sd
S9_BETA1      <- 0.8     # amount-channel effect size (H_amount / H_both controls)
S9_BROKEN_FRAC<- 0.5     # fraction of correlations shrunk in Sigma_broken
S9_CTRL_N     <- 200L    # n for the H_amount / H_both control scenarios
S9_DO_GLASSO  <- FALSE   # also run lioness_glasso in the recovery block (slow:
                         # n glasso solves per simulated study)
```


## Calibration: everything copied from the real Elbe unit

```{r s9-calibrate}
set.seed(S9_SEED)

elbe_sites <- sites_meta$site_id[sites_meta$river_basin == "Elbe"]
pu   <- prep_unit("conc", elbe_sites)
Xcal <- pu$Xlog                     # 153 x p, log10, MDL/2-floored -- the real thing
p    <- ncol(Xcal)

MU    <- colMeans(Xcal)
SDV   <- apply(Xcal, 2, sd)
SIGMA <- cor(Xcal)                  # the basin's real co-occurrence structure
FLOOR <- pu$pseudo[colnames(Xcal)]  # MDL/2 per compound (ng/L), prep_unit's floor
MDLV  <- 2 * FLOOR                  # floor = MDL/2  =>  MDL = 2 * floor

sum_tu_all <- if (ncol(tu_mat) > 0) colSums(tu_mat) else setNames(rep(1, ncol(conc_mat)), colnames(conc_mat))
TU_POOL <- sum_tu_all[elbe_sites]   # real stress axis to resample from
## [S9-fix 17.09.2026] 11 Elbe sites have no TU column (no EC10 coverage) ->
## their sum_tu is NA; resampling must draw only from sites with a real value.
TU_POOL <- TU_POOL[!is.na(TU_POOL)]
message(sprintf("[S9] TU pool: %d of %d Elbe sites have a sum-TU value",
                length(TU_POOL), length(elbe_sites)))

## source direction: first PC of the real correlation matrix. The mean shift
## alpha * z(stress) * g makes stressed sites more polluted ALONG the basin's
## real co-occurrence gradient, not uniformly.
.eig <- eigen(SIGMA, symmetric = TRUE)
G_PC <- .eig$vectors[, 1]
G_PC <- G_PC * sign(sum(G_PC))      # sign convention: mostly positive loadings

## @KEY the broken structure: half of the real off-diagonal correlations
## shrunk to 10 % of their value, then repaired to the nearest valid
## correlation matrix (eigenvalue flooring). Fixed once per run, so every
## simulated site deviates from the SAME intact structure.
.s9_near_cor <- function(M, eps = 1e-4) {
  M <- (M + t(M)) / 2
  e <- eigen(M, symmetric = TRUE)
  w <- pmax(e$values, eps)
  M2 <- e$vectors %*% (t(e$vectors) * w)   # E diag(w) E'
  d <- sqrt(diag(M2))
  M2 <- M2 / outer(d, d)
  (M2 + t(M2)) / 2
}
sel <- matrix(FALSE, p, p)
sel[upper.tri(sel)] <- sample(c(TRUE, FALSE), p * (p - 1) / 2,
                              replace = TRUE, prob = c(S9_BROKEN_FRAC, 1 - S9_BROKEN_FRAC))
sel <- sel | t(sel)
SIGMA_B <- SIGMA
SIGMA_B[sel] <- SIGMA_B[sel] * 0.1
SIGMA_B <- .s9_near_cor(SIGMA_B)

## matrix square roots (eigen factors) for fast draws:
## x = sqrt(1-k) * L %*% z1 + sqrt(k) * LB %*% z2 has exactly
## cov = (1-k) * Sigma + k * Sigma_broken -- no per-site eigendecomposition.
.s9_factor <- function(Sigma, sdv) {
  C <- diag(sdv) %*% Sigma %*% diag(sdv)
  e <- eigen(C, symmetric = TRUE)
  e$vectors %*% diag(sqrt(pmax(e$values, 0)))
}
L_INTACT <- .s9_factor(SIGMA, SDV)
L_BROKEN <- .s9_factor(SIGMA_B, SDV)

## skeleton the strength is summed over: the REAL fitted Elbe glasso skeleton
## when cached (the pipeline's actual edge set), else a top-|r| fallback at
## 5 % density. Fixed across all sims -- it plays the role of the basin
## aggregate skeleton in S6/S7.
fit_elbe <- load_fitted("conc", "Basin_Elbe")
if (!is.null(fit_elbe) && all(colnames(Xcal) %in% rownames(fit_elbe$A_glasso))) {
  SKEL <- fit_elbe$A_glasso[colnames(Xcal), colnames(Xcal)] != 0
  skel_src <- "cached Elbe glasso fit"
} else {
  thr <- quantile(abs(SIGMA[upper.tri(SIGMA)]), 0.95)
  SKEL <- abs(SIGMA) > thr; diag(SKEL) <- FALSE
  skel_src <- "top-5% |r| fallback (no cached fit)"
}
SKEL_NUM <- SKEL * 1
dimnames(SKEL_NUM) <- dimnames(SIGMA)
message(sprintf("[S9] calibrated on Elbe: %d sites x %d compounds, skeleton %d edges (%s)",
                nrow(Xcal), p, sum(SKEL) / 2, skel_src))
```


## The simulators

```{r s9-simulators}
## ---- chemistry -------------------------------------------------------------
## n sites: stress resampled from the real TU pool; kappa linked to stress on
## the logit scale plus noise; latent log-concentrations drawn from the
## kappa-mixture of the intact and broken structures; then the real MDL
## censoring + MDL/2 floor + log10 (exactly prep_unit's damage).
s9_sim_chemistry <- function(n) {
  s  <- sample(TU_POOL, n, replace = TRUE)
  zs <- as.numeric(scale(log10(s)))
  kappa <- plogis(S9_BETA_KAPPA * zs + rnorm(n, 0, S9_KAPPA_NOISE))

  Z1 <- matrix(rnorm(n * p), n, p) %*% t(L_INTACT)
  Z2 <- matrix(rnorm(n * p), n, p) %*% t(L_BROKEN)
  X  <- Z1 * sqrt(1 - kappa) + Z2 * sqrt(kappa)      # kappa recycles per row
  X  <- sweep(X, 2, MU, `+`)
  X  <- X + S9_ALPHA_MEAN * (zs %o% G_PC)            # mean shift along PC1

  C <- 10^X
  below <- C < matrix(MDLV, n, p, byrow = TRUE)
  C[below] <- matrix(FLOOR, n, p, byrow = TRUE)[below]
  Xf <- log10(C)
  colnames(Xf) <- colnames(Xcal)
  rownames(Xf) <- paste0("sim_", seq_len(n))
  list(X = Xf, kappa = kappa, tu = s)
}

## ---- microbiome --------------------------------------------------------------
## Dirichlet-multinomial counts; 20 % sensitive / 10 % tolerant / 70 % neutral
## taxa; expected abundances shift with z(sum-TU) (amount channel, beta1) and
## z(kappa) (structure channel, beta2). Metric = relative abundance of the
## sensitive class (a SPEAR-like index).
s9_sim_microbiome <- function(kappa, tu, beta1, beta2) {
  n  <- length(kappa)
  zt <- as.numeric(scale(log10(tu)))
  zk <- as.numeric(scale(kappa))
  n_sens <- round(0.2 * S9_NTAXA); n_tol <- round(0.1 * S9_NTAXA)
  sens <- c(rep(-1, n_sens), rep(1, n_tol), rep(0, S9_NTAXA - n_sens - n_tol))
  base <- rlnorm(S9_NTAXA, 0, 1.2)

  eta <- outer(zt, beta1 * sens) + outer(zk, beta2 * sens)
  eta <- sweep(eta, 2, log(base), `+`)
  alpha <- exp(eta) * S9_DM_PHI

  Gm <- matrix(rgamma(n * S9_NTAXA, shape = as.vector(alpha), rate = 1), n, S9_NTAXA)
  P  <- Gm / rowSums(Gm)
  counts <- t(vapply(seq_len(n), function(q) rmultinom(1, S9_DEPTH, P[q, ]),
                     numeric(S9_NTAXA)))
  metric <- rowSums(counts[, sens == -1, drop = FALSE]) / S9_DEPTH
  list(counts = counts, metric = metric)
}

## ---- recovery: run the unmodified S6 estimators on fake chemistry ------------
## rank-based AUC (Mann-Whitney), no extra packages
.s9_auc <- function(score, truth01) {
  r <- rank(score); m <- sum(truth01 == 1L); nn <- sum(truth01 == 0L)
  (sum(r[truth01 == 1L]) - m * (m + 1) / 2) / (m * nn)
}

s9_recovery <- function(X, kappa) {
  atyp <- as.integer(kappa > quantile(kappa, 0.75))   # top quartile = truly atypical
  bo <- bonobo_cor(X, progress = FALSE, compute_sparse = FALSE)
  li <- lioness_cor(X, progress = FALSE)
  sb <- single_site_summary(bo$nets, A_agg = SKEL_NUM)$strength_skeleton
  sl <- single_site_summary(li$nets, A_agg = SKEL_NUM)$strength_skeleton
  out <- data.frame(
    rho_bonobo   = cor(sb, kappa, method = "spearman"),
    rho_lioness  = cor(sl, kappa, method = "spearman"),
    auc_bonobo   = .s9_auc(-sb, atyp),     # LOW strength should flag atypical
    auc_lioness  = .s9_auc(-sl, atyp))
  if (S9_DO_GLASSO && !is.null(fit_elbe)) {
    lg <- lioness_glasso(X, lambda = fit_elbe$lambda_glasso, progress = FALSE)
    sg <- single_site_summary(lg$nets, A_agg = SKEL_NUM)$strength_skeleton
    out$rho_glasso <- cor(sg, kappa, method = "spearman")
    out$auc_glasso <- .s9_auc(-sg, atyp)
  }
  out
}

## ---- the S7 biomarker test, replayed on simulated data -----------------------
s9_biomarker <- function(metric, tu, strength) {
  d <- data.frame(metric = metric,
                  zt = as.numeric(scale(log10(tu))),
                  zs = as.numeric(scale(strength)))
  anova(lm(metric ~ zt, data = d), lm(metric ~ zt + zs, data = d))$`Pr(>F)`[2]
}

## one simulated study: chemistry -> BONOBO strength -> microbiome -> F-test
s9_one_study <- function(n, beta1, beta2) {
  sim <- s9_sim_chemistry(n)
  bo  <- bonobo_cor(sim$X, progress = FALSE, compute_sparse = FALSE)
  sb  <- single_site_summary(bo$nets, A_agg = SKEL_NUM)$strength_skeleton
  mb  <- s9_sim_microbiome(sim$kappa, sim$tu, beta1 = beta1, beta2 = beta2)
  s9_biomarker(mb$metric, sim$tu, sb)
}
```


## Recovery: do the estimators find the truly decoupled sites?

```{r s9-recovery}
## one example study at n = 200 for the scatter, then the n-grid for AUC/rho.
set.seed(S9_SEED + 1L)
ex  <- s9_sim_chemistry(200L)
ex_bo <- bonobo_cor(ex$X, progress = FALSE, compute_sparse = FALSE)
ex_li <- lioness_cor(ex$X, progress = FALSE)
ex_sb <- single_site_summary(ex_bo$nets, A_agg = SKEL_NUM)$strength_skeleton
ex_sl <- single_site_summary(ex_li$nets, A_agg = SKEL_NUM)$strength_skeleton

rec_rows <- list()
for (n in S9_GRID_N) {
  for (rep in seq_len(S9_NREP)) {
    set.seed(S9_SEED + 100L * n + rep)
    sim <- s9_sim_chemistry(n)
    rec_rows[[length(rec_rows) + 1]] <- cbind(n = n, rep = rep, s9_recovery(sim$X, sim$kappa))
  }
  message(sprintf("[S9] recovery grid: n = %d done", n))
}
recovery <- do.call(rbind, rec_rows)
write.csv(recovery, file.path(outroot, "tableS9_recovery.csv"), row.names = FALSE)

rec_summ <- aggregate(. ~ n, data = recovery[, c("n", "rho_bonobo", "rho_lioness",
                                                 "auc_bonobo", "auc_lioness")], mean)
print(rec_summ, row.names = FALSE)
```


## The power grid: biomarker F-test over n x beta2 (H_structure column: beta1 = 0)

```{r s9-power}
grid <- expand.grid(n = S9_GRID_N, beta2 = S9_GRID_BETA2)
pw_rows <- list()
for (i in seq_len(nrow(grid))) {
  n <- grid$n[i]; b2 <- grid$beta2[i]
  pv <- vapply(seq_len(S9_NREP), function(rep) {
    set.seed(S9_SEED + 100000L * i + rep)
    s9_one_study(n, beta1 = 0, beta2 = b2)
  }, numeric(1))
  pw_rows[[i]] <- data.frame(n = n, beta2 = b2, power = mean(pv < 0.05),
                             median_p = median(pv))
  message(sprintf("[S9] power grid: n = %d, beta2 = %.1f -> power %.2f", n, b2, mean(pv < 0.05)))
}
power_tab <- do.call(rbind, pw_rows)
write.csv(power_tab, file.path(outroot, "tableS9_power.csv"), row.names = FALSE)
print(power_tab, row.names = FALSE)
```


## Control scenarios: H_amount (specificity) and H_both (added value beyond TU)

```{r s9-controls}
## H_amount: biology responds to sum-TU only. The strength term should NOT
## survive the F-test often (specificity: rejection rate ~ 0.05).
pv_amount <- vapply(seq_len(S9_NREP), function(rep) {
  set.seed(S9_SEED + 900000L + rep)
  s9_one_study(S9_CTRL_N, beta1 = S9_BETA1, beta2 = 0)
}, numeric(1))

## H_both: both channels active. The strength term SHOULD win -- this is the
## realistic version of "strength explains biology beyond sum-TU".
pv_both <- vapply(seq_len(S9_NREP), function(rep) {
  set.seed(S9_SEED + 950000L + rep)
  s9_one_study(S9_CTRL_N, beta1 = S9_BETA1, beta2 = 0.6)
}, numeric(1))

controls <- data.frame(
  scenario      = c("H0 (beta2 = 0 in grid above)", "H_amount (beta1 = 0.8 only)",
                    "H_both (beta1 = 0.8, beta2 = 0.6)"),
  rejection_rate = c(power_tab$power[power_tab$n == S9_CTRL_N & power_tab$beta2 == 0],
                     mean(pv_amount < 0.05), mean(pv_both < 0.05)))
write.csv(controls, file.path(outroot, "tableS9_controls.csv"), row.names = FALSE)
print(controls, row.names = FALSE)
```


## Figures

```{r s9-figures}
png(file.path(outroot, "figS9_recovery.png"), width = 13, height = 6.5,
    units = "in", res = 150)
par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

## (a) one example study: strength vs TRUE kappa, both estimators
plot(ex$kappa, ex_sb, pch = 16, cex = 0.7, col = "#2A9D8F",
     xlab = expression("true decoupling " * kappa[q]),
     ylab = "strength on skeleton", main = "(a) recovery, one study (n = 200)")
.rs <- function(v, to) to[1] + (v - min(v)) / (max(v) - min(v)) * diff(to)
points(ex$kappa, .rs(ex_sl, range(ex_sb)), pch = 16, cex = 0.7, col = "#E63946")
legend("topright", bty = "n", pch = 16, col = c("#2A9D8F", "#E63946"),
       legend = c(sprintf("BONOBO  rho = %.2f", cor(ex_sb, ex$kappa, method = "spearman")),
                  sprintf("LIONESS rho = %.2f", cor(ex_sl, ex$kappa, method = "spearman"))))

## (b) AUC against the true atypical labels, per estimator per n
plot(NA, xlim = range(S9_GRID_N), ylim = c(0.5, 1), xlab = "sites per study (n)",
     ylab = "AUC (top-quartile kappa)", main = "(b) atypical-site detection", log = "x")
for (m in c("auc_bonobo", "auc_lioness")) {
  mm <- aggregate(recovery[[m]], by = list(n = recovery$n), mean)
  se <- aggregate(recovery[[m]], by = list(n = recovery$n),
                  function(v) sd(v) / sqrt(length(v)))
  lines(mm$n, mm$x, col = if (m == "auc_bonobo") "#2A9D8F" else "#E63946", lwd = 2)
  points(mm$n, mm$x, pch = 16, col = if (m == "auc_bonobo") "#2A9D8F" else "#E63946")
  segments(mm$n, mm$x - se$x, mm$n, mm$x + se$x,
           col = if (m == "auc_bonobo") "#2A9D8F" else "#E63946")
}
abline(h = 0.5, lty = 3, col = "grey60")
legend("bottomright", bty = "n", lwd = 2, col = c("#2A9D8F", "#E63946"),
       legend = c("BONOBO", "LIONESS"))
dev.off()

png(file.path(outroot, "figS9_power.png"), width = 8, height = 6.5,
    units = "in", res = 150)
par(mar = c(4.5, 4.5, 3, 1))
cols <- colorRampPalette(c("#E9C46A", "#BC6C25"))(length(S9_GRID_BETA2))
plot(NA, xlim = range(S9_GRID_N), ylim = c(0, 1), xlab = "sites per study (n)",
     ylab = "power (F-test p < 0.05)", log = "x",
     main = "detection power for a structure-driven biological response")
for (j in seq_along(S9_GRID_BETA2)) {
  sub <- power_tab[power_tab$beta2 == S9_GRID_BETA2[j], ]
  lines(sub$n, sub$power, col = cols[j], lwd = 2,
        lty = if (S9_GRID_BETA2[j] == 0) 3 else 1)
  points(sub$n, sub$power, pch = 16, col = cols[j])
}
abline(h = c(0.05, 0.8), lty = 3, col = "grey60")
legend("bottomright", bty = "n", lwd = 2, col = cols,
       legend = paste0("beta2 = ", S9_GRID_BETA2), title = "structure effect")
dev.off()

message("[S9] figures written: figS9_recovery.png, figS9_power.png")
```


## How to read the output

* **Recovery table** (`tableS9_recovery.csv`, `figS9_recovery.png`): if the
  S7/S8 interpretation is right, BONOBO strength should correlate NEGATIVELY
  with the true kappa (rho clearly below 0) and flag the truly-decoupled sites
  (AUC well above 0.5), while LIONESS — which conflates amount with pattern —
  should do visibly worse, because in this simulation amount (the PC1 mean
  shift) and decoupling both grow with stress.
* **Power table** (`tableS9_power.csv`, `figS9_power.png`): the beta2 = 0 row
  is the H0 false-positive rate and should sit near 0.05. Read the table as:
  "a same-site chemical + eDNA study with n = X sites detects a structure
  effect of size beta2 = Y with the tabulated power." The n where the
  beta2 = 0.6 line crosses 0.8 is the recommended minimum size of the
  follow-up study.
* **Controls** (`tableS9_controls.csv`): H_amount must stay near 0.05 —
  otherwise the biomarker test is not specific and the S7 null result would
  need reinterpretation. H_both shows the test does detect added value of
  strength when both channels are real.
* **Caveats** (state them in the thesis): this is an identifiability study.
  The Dirichlet-multinomial community is a stand-in for real eDNA data;
  kappa's link to stress is assumed, not measured; and the simulation cannot
  say whether structure-driven biology EXISTS — only that the pipeline would
  find it at the tabulated sample sizes if it did.


# Session info

```{r session-info}
sessionInfo()
```
